---
title: "reviewtable_organize"
author: "tksym"
date: today
params:
  in1: null
  in2: null
  dir_out: null
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    fig-format: png
    embed-resources: true
    self-contained-math: true
    code-fold: true
    code-overflow: wrap
  pdf:
    toc: false
editor: 
  mode: source
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 16,
  fig.height = 9,
  dpi = 300,
  out.width = "100%",
  fig.align = "center",
  cache = FALSE
)

time_start <- Sys.time()
```

```{r}
#| label: packages

pacman::p_load(
  # data import
  here,
  fs,

  # data manipulation
  tidyverse,
  
  # plot
  ggh4x,
  patchwork,
  gt,

  # misc
  tictoc,
  sessioninfo,
  report
)

```

```{r}
#| label: othersettings
#| include: false

# assign from outside
in1 <- params$in1
in2 <- params$in2
dir_out <- params$dir_out


# source file
rts <- "C:/Users/wisdo/Dropbox/02_proj_cloud/_localrepo/basicscripts/script"
source(here(rts, "othersettings.R"))
# source(here(rts, "gs.R"))


# 
```

# WORK FLOW

- Import data

- Check original data

- Data cleaning

- Visualize number of range shift observations per ecosystem

# DATA

Importing previous review table: `{r} fs::path_rel(in1)`

Importing updated review table: `{r} fs::path_rel(in1)`

Output directory: `{r} fs::path_rel(dir_out)`



# Import

```{r}

# original data
rt_prev <- read_csv(in1)


# data after small checks in excel
# rt <- read_csv(in2)

library(readxl)
rt <- read_xlsx(in2, sheet = "dataentry_ok")

# other sheets
lon_latrange <- read_xlsx(in2, sheet = "longitude_latrange")
ref_en <- read_xlsx(in2, sheet = "En-list_ok")
ref_ja <- read_xlsx(in2, sheet = "Ja-list_ok")




# ============================================== #
# import color code used in polygon

studyarea_color <- readr::read_csv(
  "C:/Users/wisdo/Dropbox/02_proj_cloud/_localrepo/er_review/result/create_leaflet_map/studyarea_color.csv"
)

```



# Check original data

```{r}
#| eval: FALSE
#| echo: FALSE


# glimpse(rt_prev)

# ============================================== #
# vars for simple cleaning

cols_upper_prev <- c("recorder_name",
                "reference_no",
                "study_region",
                "area_name",
                "taxonomic_unit",
                "taxonomic_unit_revised")




# ============================================== #
# simple cleaning

rt_prev2 <- rt_prev %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  
  # convert text to lower (except specified columns)
  mutate(across(c(where(is.character),
                  -all_of(cols_upper_prev)),
                str_to_lower)) %>% 
  
  # replace NA with "na"
  mutate(across(everything(),
                ~ tidyr::replace_na(.x, "na")))




# glimpse(rt_prev2)
```


```{r}
#| eval: FALSE
#| echo: FALSE

# ============================================== #
# check: show unique values for each column

nrow(rt_prev2)

for (nm in names(rt_prev2)) {
  vals <- unique(rt_prev2[[nm]])

  cat("\n====================\n")
  cat("Column:", nm, "\n")
  cat("n_unique:", length(vals), "\n")
  cat("====================\n")

  print(vals)

}


```







# Clean range shfit data

## rt2: light cleaning

```{r}
# ============================================== #
# simple cleaning

# glimpse(rt)
# glimpse(rt2)


cols_upper <- c("recorder_name",
                "reference_no",
                "study_area",
                "specific_area",
                "taxonomic_unit",
                "taxonomic_unit_revised",
                "kingdom",
                "phylum",
                "class",
                "order",
                "family",
                "genus")



rt2 <- rt %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  
  # convert text to lower (except specified columns)
  mutate(across(c(where(is.character),
                  -all_of(cols_upper)),
                str_to_lower)) %>% 
  
  # replace NA with "na"
  mutate(across(everything(),
                ~ tidyr::replace_na(.x, "na"))) %>% 
  # taxonomy
  relocate(kingdom, phylum, class, order, family, genus, 
           .after = taxonomic_unit_revised)

# glimpse(rt2)
```


## rt3: light cleaning

```{r}
# ============================================== #
# filter out irrelevant data at initial stage
# edit start period and end period data

rt2$include %>% unique() %>% sort()

rt3 <- rt2 %>% 
  filter(
    # this include col was made in excel for rows based on:
    # n_years: too large (comparison with glacial time)
    # gradient = na
    # magnitude_of_shift = na
    # notes = no resurvey
    # include == 2 is "pending" literature
    include %in% 1) %>%
  
  # start period
  rename(startperiod_orig = start_period) %>%
  mutate(
    # extract all 4-digit years
    startperiod_years = str_extract_all(startperiod_orig, "\\d{4}"),
    
    startperiod_startyear = map_int(
      startperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[1]))
    ),
    
    startperiod_endyear = map_int(
      startperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[length(.x)]))
    ),
    
    startperiod_midyear = (startperiod_startyear + startperiod_endyear) / 2
  ) %>%
  select(-startperiod_years) %>% 
  relocate(startperiod_startyear, startperiod_endyear, startperiod_midyear, .after = startperiod_orig) %>% 
  
  # end period
  rename(endperiod_orig = end_period) %>%
  mutate(
    # extract all 4-digit years
    endperiod_years = str_extract_all(endperiod_orig, "\\d{4}"),
    
    endperiod_startyear = map_int(
      endperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[1]))
    ),
    
    endperiod_endyear = map_int(
      endperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[length(.x)]))
    ),
    
    endperiod_midyear = (endperiod_startyear + endperiod_endyear) / 2
  ) %>%
  select(-endperiod_years) %>% 
  relocate(endperiod_startyear, endperiod_endyear, endperiod_midyear, .after = endperiod_orig) %>% 
  
  # n_years
  mutate(
    n_years_midyear = endperiod_midyear - startperiod_midyear
  ) %>% 
  relocate(n_years_midyear, .after = n_years) %>% 
  mutate(
  azimuth_of_shift = na_if(azimuth_of_shift, "na"),
  azimuth_of_shift = as.numeric(azimuth_of_shift)
)

nrow(rt2)
nrow(rt3)
```




## reference info

```{r}

# ============================================== #
# minor edits

# glimpse(ref_en)
# glimpse(ref_ja)

ref_en2 <- ref_en %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower())

ref_ja2 <- ref_ja %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower())


# glimpse(ref_en2)
# glimpse(ref_ja2)


ref_en3 <- ref_en2 %>% 
  filter(reference_no %in% unique(rt3$reference_no)) %>% 
  select(reference_no, article_title, data_section, source_title, publication_year, authors, doi) %>% 
  mutate(publication_year = as.numeric(publication_year),
         language = "English")

ref_ja3 <- ref_ja2 %>% 
  filter(reference_no %in% unique(rt3$reference_no)) %>% 
  select(reference_no, article_title, data_section, source_title, publication_year, authors, doi) %>% 
  mutate(language = "Japanese")

# glimpse(ref_en3)
# glimpse(ref_ja3)



# ============================================== #
# bind
ref_bind <- bind_rows(ref_en3,
                      ref_ja3) %>% 
  relocate(language, reference_no, doi, authors, article_title, publication_year, source_title, data_section) %>% 
  mutate(author_year = str_c(authors, publication_year, sep = " "))


# ============================================== #
# check and change cell entries


# ------------ #
ref_bind$author_year %>% unique()
ref_bind$author_year %>% class()

authors_recode <- c(
  "Sakio & Masuzawa 2012" = "Sakio & Masuzawa 2012",
  "Jung, S; Pang, IC; Lee, JH; Choi, I; Cha, HK 2014" = "Jung et al. 2014",
  "Winkler, DE; Amagai, Y; Huxman, TE; Kaneko, M; Kudo, G 2016" = "Winkler et al. 2016",
  "Shimazaki, M; Sasaki, T; Hikosaka, K; Nakashizuka, T 2011" = "Shimazaki et al. 2011",
  "Tougou, D; Musolin, DL; Fujisaki, K 2009" = "Tougou et al. 2009",
  "Kakehi, S; Narimatsu, Y; Okamura, Y; Yagura, A; Ito, S 2021" = "Kakehi et al. 2021",
  "Kurihara, T; Takami, H; Kosuge, T; Chiba, S; Iseda, M; Sasaki, T 2011" = "Kurihara et al. 2011",
  "T Sakiyama 2024" = "Sakiyama 2024",
  "Naoki H. Kumagai, Jorge García Molinosa, Hiroya Yamano, Shintaro Takao, Masahiko Fujii, Yasuhiro Yamanaka 2018" = "Kumagai et al. 2018",
  "S. Watari, S. Takemura, H. Oyaizu 2024" = "Watari et al. 2024",
  "Tu, CY; Tian, YJ; Hsieh, CH 2015" = "Tu et al. 2015",
  "Yamano, H; Sugihara, K; Nomura, K 2011" = "Yamano et al. 2011",
  "Y Kawauchi, Y Yagi, T Yano, K Fujiwara 2023" = "Kawauchi et al. 2023",
  "K Higashiguchi, S Nishijima, Y Kamimura… 2025" = "Higashiguchi et al. 2025",
  "Y Kawakami, K Yamazaki, K Ohashi 2014" = "Kawakami et al. 2014",
  "山内 貴義, 工藤 雅志, 高槻 成紀 2007" = "Yamauchi et al. 2007",
  "北原 正彦 2008" = "Kitahara 2008",
  "木澤 遼; 濱 侃; 吉田 圭一郎 2022" = "Kizawa et al. 2022",
  "奥村 忠誠; 清水 庸; 大政 謙次 2009" = "Okumura et al. 2009"
)


# ------------ #
ref_bind$source_title %>% unique()

journal_recode <- c(
  "Journal of Plant Research" = "Journal of Plant Research",
  "Reviews in fish biology and fisheries" = "Reviews in Fish Biology and Fisheries",
  "Plant ecology" = "Plant Ecology",
  "Global change biology" = "Global Change Biology",
  "Entomologia experimentalis et applicata" = "Entomologia Experimentalis et Applicata",
  "Marine ecology progress series" = "Marine Ecology Progress Series",
  "Marine biology" = "Marine Biology",
  "PhD dissertation" = "PhD dissertation",
  "PNAS" = "PNAS",
  "Frontiers in Marine Science" = "Frontiers in Marine Science",
  "Fisheries oceanography" = "Fisheries Oceanography",
  "Geophysical research letters" = "Geophysical Research Letters",
  "Regional studies in marine science" = "Regional Studies in Marine Science",
  "SSRN" = "SSRN",
  "Applied entomology and zoology" = "Applied Entomology and Zoology",
  "哺乳類科学" = "哺乳類科学",
  "昆虫と自然" = "昆虫と自然",
  "E-journal GEO" = "E-journal GEO",
  "環境科学会誌" = "環境科学会誌"
)


# ------------ #

ref_bind2 <- ref_bind %>% 
  mutate(publication_id = recode(author_year, !!!authors_recode)) %>% 
  relocate(publication_id, .after = reference_no) %>% 
  select(-author_year) %>%
  mutate(
    source_title = recode(source_title, !!!journal_recode)
  ) %>% 
  arrange(language, publication_year)





```



## recodes

```{r}
# ============================================== #
# column names

names(rt3)



# ============================================== #
# check and change cell entries


# ------------ #
rt3$geographic_scale %>% unique()

geographicscale_recode <- c()


# ------------ #
rt3$study_area %>% unique()

studyarea_recode <- c()


# ------------ #
rt3$specific_area %>% unique()

specificarea_recode <- c()


# ------------ #
rt3$kingdom %>% unique()

kingdom_recode <- c(
  "Plant" = "Plantae",
  "animal" = "Animalia",
  "Animal" = "Animalia",
  "Chromista" = "Chromista",
  "Animalia"  = "Animalia",
  "Plantae"  = "Plantae",
  "動物" = "Animalia",
  "植物" = "Plantae"
  
)

kingdom_recode %>% unique()

# ------------ #
rt3$taxonomic_resolution %>% unique()

taxres_recode <- c(
  "species-level" = "species",
  "species" = "species",
  "種レベル" = "species",
  "order-level" = "order")

taxres_recode %>% unique()


# ------------ #
rt3$taxonomic_unit %>% unique()

# recode taxonomic_unit_revised
taxunit_recode <- c()


# ------------ #
rt3$taxonomic_unit_revised %>% unique() %>% sort()

# supported by ai
taxunitrev_recode <- c(

  # check later: na
  "na" = "na"
)

taxunitrev_recode %>% unique()


# ------------ #
rt3$ecosystem %>% unique()


ecosystem_recode <- c(
  
  # alpine
  "alpine" = "alpine",
  
  # forest
  "forest" = "forest",
  "森林" = "forest",
  
  
  "forest, alpine" = "forest", # Hitomi E0267 Hakkoda, include subalpine as forest
  
  # grassland
  "grassland" = "grassland",
  
  # wetland
  "wetlands" = "wetland",
  
  # marine
  "marine" = "marine",
  
  
  
  # check
  
  # mixed ecosystems as terrestrial
  "forest, grassland" = "terrestrial", # Dai E0400 Nezara viridula kamemushi
  "森林、草地" = "terrestrial", # Nonaka J0202 sika
  "限定なし" ="terrestrial", # Nonaka J0201 Birds
  
  # broad terrestrial kept as terrestrial
  "terrestrial" = "terrestrial" # Sueyoshi E2247 ladybug
)

ecosystem_recode %>% unique()

realm_recode <- c(
  
  # terrestrial
  "alpine" = "terrestrial",
  "forest" = "terrestrial",
  "grassland" = "terrestrial",
  "wetland" = "terrestrial",
  
  # tentative terrestrial
  "urban" = "terrestrial",

    
  # marine
  "marine" = "marine"
)


# ------------ #
rt3$information_source %>% unique()

infosrc_recode <- c(
  
  # resurvey by authors
  "resurvey" =
    "resurvey",
  "authors did the resurvey" =
    "resurvey",
  "著者が再調査を実施" =
    "resurvey",
  
  
  # remote sensing
  "remote sensing" =
    "remote sensing",
  "aerial photograph by forest agency" =
    "remote sensing",
  "geospatial information authority of japan" =
    "remote sensing",
  
  
  # literature
  "literature" =
    "literature",
  "published papers" =
    "literature",
  
  
  # literature and speciemens
  "literature and museum specimens" =
    "literature and specimens",
  "specimen records, literature" =
    "literature and specimens",
  

  # citizen and specimens
  "citizen science and specimens" =
    "citizen science and specimens",
  
  
  
  # check later with Jorge: fishery statistics, separate trawl or not?
  "fishery statistics" =
    "fishery statistics",
  "commercial fisheries data" =
    "fishery statistics",
  "fishery and aquaculture production statistics 　国の漁獲統計" =
    "fishery statistics",
  
  "bottom-trawl surveys" =
    "trawl survey",
  "trawl fishery data" =
    "trawl fishery",
  "fishery-independent trawl surveys by fisheries research and education agency (fra)" =
    "fishery-independent trawl survey",
  
  
  # check later: governmental survey
  "distribution survey by the ministry of the environment" =
    "governmental monitoring survey",
  
  
  # check later: database
  "データベースを利用" =
    "multiple sources",
  
  # check 
  "na" =
    "na" # Fatima E0608
)

infosrc_recode %>% unique()


# ------------ #
rt3$gradient %>% unique() %>% sort()

gradient_recode <- c(
  
  # elevation
  "elevation" = "elevation",
  "標高" = "elevation",
  
  # lat
  "latitude" = "latitude",
  "latitude (northward)" = "latitude",
  "latitude (poleward / northward)" = "latitude",
  
  
  # lon
  "longitude" = "longitude",
  
  
  # horizontal
  "horizontal" = "horizontal", # Jorge: I think this is ok
  "水平方向" = "horizontal", # Nonaka J0202
  "緯度・経度方向" = "horizontal" # Nonaka J0201 sika: checked and this is based on mesh data so this is horizontal
  )

gradient_recode %>% unique()



# ------------ #
rt3$position %>% unique()

position_recode <- c(
  
  # leading edge
  "leading edge" = "leading edge",
  "leading" = "leading edge",
  "northern limit" = "leading edge",
  "前縁" = "leading edge",
  
  # centroid
  "centroid" = "centroid",
  "centroid / multiple sites" = "centroid",
  
  "average" = "centroid", # Dai E0434 pika: study is about centroid
  "na" = "centroid", # Nonaka J0202 sika: authors do not mention centroid, but the study area covers entire honshu, shikoku, and kyushu, which largely covers the sika's distribution. so centroid was applied here.
  
  # need to check using distribution data, but it will be converted to centroid for now, because it is likely to reflect the centroid
  "unknown" = "centroid", # Sakiyama E0009
  
  # trailing edge
  "trailing edge" = "trailing edge",
  "trailing" = "trailing edge",
  
  
  # check
  "前縁、重心、後縁" = "others" # Nonaka J0201 bird: authors do not mention but this can be interpreted as centroid


)

position_recode %>% unique()



# ------------ #
rt3$measurement_type %>% unique()

meastype_recode <- c(
  
  # abundance
  "abundance" = "abundance",
  
  # occurrence
  "occurrence" = "occurrence",
  "在不在" = "occurrence",

  # presence-only: group as occurrence
  "occurrence (presence-only)" = "occurrence_presonly",
  
  
  
  # check
  
  # density: group as abundance
  "density from catch data" = "abundance_density",
  "density from survey" = "abundance_density",
  
  # cpue: group as abundance
  "cpue" = "abundance_cpue",
  "abundance (cpue)" = "abundance_cpue",
  
  # catch: group as abundance
  "catch landings" = "abundance_catchlandings"
  )

meastype_recode %>% unique()


observation_recode <- c(
  
  # abundance
  "abundance" = "abundance",
  "abundance_density" = "abundance",
  "abundance_cpue" = "abundance",
  "abundance_catchlandings" = "abundance",
  
  # occurrence
  "occurrence" = "occurrence",
  "occurrence_presonly" = "occurrence"
  
  
)




# ------------ #
rt3$startperiod_orig %>% unique()



# ------------ #
rt3$endperiod_orig %>% unique()



# ------------ #
rt3$n_years %>% unique()


# ------------ #
rt3$n_years_midyear %>% unique()



# ------------ #
rt3$number_of_censuses %>% unique()

ncensus_recode <- c(
  
  # check
  "na" = "na", # Harsh E0949, E1034, E_har1; Jorge E_jgm1, E1585, E1667, E1991; Fatima E0608
  
  "a total of 632 study sites over 16 years (26-59 sites per year)" = "632", # Sueyoshi E2241
  
  "the number of specimens recorded was 1312, but the number obtained from the literature was not reported." = "1312" # Sueyoshi E2247
)

ncensus_recode %>% unique()


# ------------ #
rt3$magnitude_of_shift %>% unique()

# magshift_recode <- c(
# )
# 
# magshift_recode %>% unique()


# ------------ #
rt3$unit_of_shift %>% unique()

# later
# convert cells with m, km, km/decade to km/year

unitshift_recode <- c(
  
  # meter
  "meter" = "m",
  "m" = "m",
  "メートル" = "m",
  
  # km
  "kilometer" = "km",
  "km" = "km",
  
  # km/year
  "km year-1" = "km/year",
  "km/year" = "km/year",
  
  # km/decade
  "km per decade" = "km/decade",
  
  # degree
  "degree e" = "degree",
  "degree n" = "degree",
  "degrees" = "degree",
  
  # degree/year
  "degree year-1" = "degree/year",
  "degree/year" = "degree/year",
  "degrees/year" = "degree/year"
)





# ------------ #
rt3$azimuth_of_shift %>% unique() %>% sort()

# azimuth_recode <- c(
#   
# )



# ------------ #
rt3$shift_tested %>% unique() %>% sort()

shifttest_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (centroid position)" = "yes",
  "estimated" = "yes", # Jorge E_jgm2, E1991: tested shift based on significance of coef

  # no
  "no" = "no")

# ------------ #
rt3$shift_significant %>% unique() %>% sort()

shiftsign_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (supported by consistent results across 6 modeling thresholds, here the magnitude is given based on equate entropy of thresholded and original distribution)" = "yes",

  # no
  "no" = "no",
  
  # na
  "na" = "na",
  "no analysis" = "na",
  "no analyses" = "na"
  
)

# ------------ #
rt3$data_cleaning %>% unique() %>% sort()

dataclean_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (the oldest records from each of the 39 regions were used)" =
    "yes",

  # no
  "no" = "no",
  
  # na
  "na" = "na"
  
  )

dataclean_recode %>% unique()


# ------------ #
rt3$modelling_conducted %>% unique() %>% sort()

# check again

model_recode <- c(
  
  # yes
  "yes" = "yes",
  
  "yes\r\n  (centroid shifts were estimated using vector autoregressive spatiotemporal model)" =
    "yes",
  
  # no
  "no" = "no"
  )

model_recode %>% unique()


# ------------ #
rt3$other_effects %>% unique() %>% sort()

othereff_recode <- c(
  
  # no
  "no" = "no",
  
  # na
  "na" = "na",
  
  # try to phrase in keywords
  
  "authors argue that the natural recovery after volcano eruption is the main process of the advancing elevation and discuss that climate change may accelerate this process." =
    "volcanic recovery",
  
  "human activities, species interactions, landuse" =
    "human activity, species interaction, land use",
  
  
  "①increase in population size ②decrease in snowfall due to global warming" =
    "population increase, snowfall decrease",
  
  "distance from moorland" =
    "distance from moorland",
  
  "species interactions" =
    "species interaction",
  
  "fishery pressure, species interactions" =
    "fishery pressure, species interaction",
  
  "sea current, species interaction, ultraviolet radiation, salinity stress" =
    "sea current, species interaction, ultraviolet radiation, salinity",
  
  "coastal development, deterioration of water quality" =
    "coastal development, water quality",
  
  "pacific decadal oscillation" =
    "pacific decadal oscillation",
  
  "adult dispersal ability, larval feeding range, distribution of food resources" =
    "dispersal ability, larval feeding, food resources",
  
  "land cover changes (increase in forest area due to abandonment of agricultural land)" =
    "land cover change",
  
  "precipitation, elevation, human influence index" =
    "precipitation, elevation, human influence",
  
  "土地利用変化、農地環境変化" =
    "land use change, agricultural change",
  
  "標高、広葉樹林面積、針葉樹林面積、草地面積、田んぼ面積、荒地面積、耕作放棄地面積、人口、道路延長距離、距離、積雪" =
    "elevation, land cover, population, road length, distance, snow"
)



# ------------ #
rt3$shift_source %>% unique() %>% sort()

shiftsrc_recode <- c(
  
  # stated
  "stated" = "stated",
  "stated in discussion" = "stated",
  "yes" = "stated", # 0434: case of northern pika

  # estimated
  "estimated" = "estimated"
  )

shiftsrc_recode %>% unique()


# ------------ #
rt3$notes %>% unique()

# notes_recode <- c()
```


## rt4: heavy recoding

```{r}
# ============================================== #
# clean dataframe by recoding

rt4 <- rt3 %>% 
  
  mutate(
  
  # ------------------------------------------------------------------ #
    # recode categorical columns
    
    # study_region = recode(study_region, !!!studyregion_recode),
    # area_name = recode(area_name, !!!areaname_recode),
    
    kingdom = recode(kingdom, !!!kingdom_recode),
    
    taxonomic_resolution = recode(
      taxonomic_resolution,
      !!!taxres_recode
    ),
    
    # taxonomic_unit = recode(
    #   taxonomic_unit,
    #   !!!taxunit_recode
    # ),
    
    taxonomic_unit_revised = recode(
      taxonomic_unit_revised,
      !!!taxunitrev_recode
    ),
    
    ecosystem = recode(
      ecosystem,
      !!!ecosystem_recode
    ),
    
    realm = recode(
      ecosystem,
      !!!realm_recode
    ),
    
    information_source = recode(
      information_source,
      !!!infosrc_recode
    ),
    
    gradient = recode(
      gradient,
      !!!gradient_recode
    ),
    
    position = recode(
      position,
      !!!position_recode
    ),
    
    measurement_type = recode(
      measurement_type,
      !!!meastype_recode
    ),
    
    observation = recode(
      measurement_type,
      !!!observation_recode
    ),
    
    # n_years = recode(
    #   n_years,
    #   !!!nyears_recode
    # ),
    
    number_of_censuses = recode(
      number_of_censuses,
      !!!ncensus_recode
    ),
    
    # magnitude_of_shift = recode(
    #   magnitude_of_shift,
    #   !!!magshift_recode
    # ),
    
    unit_of_shift = recode(
      unit_of_shift,
      !!!unitshift_recode
    ),
    
    # azimuth_of_shift = recode(
    #   azimuth_of_shift,
    #   !!!azimuth_recode
    # ),
    
    shift_tested = recode(
      shift_tested,
      !!!shifttest_recode
    ),
    
    shift_significant = recode(
      shift_significant,
      !!!shiftsign_recode
    ),
    
    data_cleaning = recode(
      data_cleaning,
      !!!dataclean_recode
    ),
    
    modelling_conducted = recode(
      modelling_conducted,
      !!!model_recode
    ),
    
    other_effects = recode(
      other_effects,
      !!!othereff_recode
    ),
    
    shift_source = recode(
      shift_source,
      !!!shiftsrc_recode
    ),
    
    # notes = recode(
    #   notes,
    #   !!!notes_recode
    # ),
    
    
    # ------------------------------------------------------------------ #
    # numeric conversion
    
    number_of_censuses = na_if(number_of_censuses, "na"),
    number_of_censuses = as.numeric(number_of_censuses),
    
    magnitude_of_shift = na_if(magnitude_of_shift, "na"),
    magnitude_of_shift = as.numeric(magnitude_of_shift)
    
  ) %>% 
  relocate(realm, .after = "ecosystem") %>% 
  relocate(observation, .after = "measurement_type") %>% 
  
  # ecosystem order
  mutate(
    
    # check entry with ecosystem_recode %>% unique()
    
    ecosystem = factor(
      ecosystem,
      levels = c(
        "alpine",
        "forest",
        "grassland",
        "wetland",
        "terrestrial",
        "marine"
      )
    ),
    realm = factor(
      realm,
      levels = c(
        "terrestrial",
        "marine"
        )
    ),
    gradient = factor(
      gradient,
      levels = c(
        "elevation",
        "latitude",
        "longitude",
        "horizontal")
    )
  ) %>% 
  
  left_join(ref_bind2 %>% select(-language),
            by = "reference_no")
```


## rt5: edit unit

```{r}
# ============================================== #
# finally edit unit


# ------------ #
# check combination of gradient and unit
rt4 %>%
  distinct(gradient, unit_of_shift) %>%
  arrange(gradient, unit_of_shift)




# ------------ #
# conversion from latitude degrees to km
latdeg_to_km <- function(deg) {
  earth_radius <- 6371
  deg * 2 * pi * earth_radius / 360
}

# ---- #
# example 

latdeg_to_km(0.1)
# ---- #



# ------------ #
# conversion from longitude degrees to km
londeg_to_km <- function(deg, latitude) {
  earth_radius <- 6371
  deg * 2 * pi * earth_radius / 360 * cos(latitude * pi / 180)
}

# ---- #
# example

londeg_to_km(1, latitude = 36)
londeg_to_km(1, latitude = 46)
# ---- #

# lat mean for longitude studies
lon_latrange2 <- lon_latrange %>% 
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  # convert text to lower (except specified columns)
  mutate(gradient = str_to_lower(gradient)) %>% 
  filter(include == 1) %>% 
  rowwise() %>% 
  mutate(lat_mean = mean(lat_min, lat_max)) %>% 
  select(reference_no, gradient, lat_mean)
  






# ------------ #
rt5 <- rt4 %>% 
  
  left_join(lon_latrange2, 
            by = c("reference_no", "gradient")) %>% 
  
  # transform magnitudeshift and unitshift
  mutate(
    
    magnitude_of_shift = case_when(
      
      # apply these to magnitudeshift depending on
      # gradient
      # unitshift
      
      
      # ------------------------------------------------------
      # elevation --> m/year
      # ------------------------------------------------------
      
      # m --> m/year
      gradient == "elevation" & unit_of_shift == "m" ~
        magnitude_of_shift / n_years_midyear,
      
      
      # ------------------------------------------------------
      # latitude --> km/year
      # ------------------------------------------------------
      
      # degree --> km --> km/year
      gradient == "latitude" & unit_of_shift == "degree" ~
        latdeg_to_km(magnitude_of_shift) / n_years_midyear,
      
      # degree/year --> km/year
      gradient == "latitude" & unit_of_shift == "degree/year" ~
        latdeg_to_km(magnitude_of_shift),
      
      # km --> km/year
      gradient == "latitude" & unit_of_shift == "km" ~
        magnitude_of_shift / n_years_midyear,
      
      # already km/year
      gradient == "latitude" & unit_of_shift == "km/year" ~
        magnitude_of_shift,
      
      
      # ------------------------------------------------------
      # longitude --> km/year
      # ------------------------------------------------------
      
      # degree --> km --> km/year
      gradient == "longitude" & unit_of_shift == "degree" ~
        londeg_to_km(magnitude_of_shift, latitude = lat_mean) / n_years_midyear,
      
      # degree/year --> km/year
      gradient == "longitude" & unit_of_shift == "degree/year" ~
        londeg_to_km(magnitude_of_shift, latitude = lat_mean),
      
      
      
      # ------------------------------------------------------
      # horizontal --> km/year
      # ------------------------------------------------------
      
      # km --> km/year
      gradient == "horizontal" & unit_of_shift == "km" ~
        magnitude_of_shift / n_years_midyear,
      
      # already km/year
      gradient == "horizontal" & unit_of_shift == "km/year" ~
        magnitude_of_shift,
      
      
      # anything unexpected
      TRUE ~ NA_real_

    ),
    
    unit_of_shift = case_when(
      gradient == "elevation" ~ "m/year",
      gradient %in% c("latitude", "longitude", "horizontal") ~ "km/year",
      TRUE ~ NA_character_
      )
    )
```


## rt6: create horizontal

```{r}
# ------------ #
# check combination of gradient and unit
rt5 %>%
  distinct(gradient, unit_of_shift) %>%
  arrange(gradient, unit_of_shift)




# ------------ #
# check distinct combinations of publication, gradient, and unit
rt5 %>%
  distinct(
    publication_id,
    gradient,
    unit_of_shift
  ) %>%
  arrange(
    publication_id,
    gradient,
    unit_of_shift
  )


rt5_latlon <- rt5 %>%
  
  group_by(
    publication_id,
    taxonomic_unit_revised,
    position,
    across(measurement_type:number_of_censuses)
  ) %>%
  
  # keep groups containing both latitude and longitude
  filter(
    all(c("latitude", "longitude") %in% gradient)
  ) %>%
  
  ungroup()


rt5_latlon_wide <- rt5_latlon %>%
  
  select(
    -c(comment:recorder_name,
       azimuth_of_shift, 
       shift_tested:notes,
       lat_mean)
  ) %>%
  
  pivot_wider(
    names_from = gradient,
    values_from = magnitude_of_shift
  ) %>%
  
  mutate(
    magnitude_of_shift = sqrt(
      longitude^2 + latitude^2
    ),
    
    azimuth_of_shift = (
      atan2(longitude, latitude) * 180 / pi + 360
    ) %% 360,
    gradient = "horizontal",
    unit_of_shift = "km/year",
    comment = "magnitude and azimuth of the horizontal shift estimated by combining latitude and longitude shift observations"
  )



rt6 <- bind_rows(rt5,
                 rt5_latlon_wide %>% 
                   select(-latitude, -longitude))




```


## final edit

```{r}

rt7 <- rt6 %>% 
  mutate(
    
    # ---------------------------------------------- #
    # based on publication
    
    ecosystem = case_when(
      publication_id == "Sakiyama 2024" ~ "forest, alpine",
      publication_id == "Shimazaki et al. 2011" ~ "subalpine",
      publication_id == "Kizawa et al. 2022" ~ "subalpine",
      
      # keep original value otherwise
      TRUE ~ as.character(ecosystem)
    ),
    
    information_source = case_when(
      information_source == "governmental monitoring survey" ~ "scientific survey",
      
      information_source == "literature" ~ "literature and specimens",
      information_source == "literature and specimens" ~ "literature and specimens",
      
      publication_id == "Kawauchi et al. 2023" ~ "fishery statistics",
      publication_id == "Tu et al. 2015" ~ "fishery statistics",
      
      publication_id == "Higashiguchi et al. 2025" ~ "scientific survey",
      publication_id == "Kakehi et al. 2021" ~ "scientific survey",
      
      # keep original value otherwise
      TRUE ~ as.character(information_source)
    )
  )


```




## export

```{r}

# ------------ #
rt_reorg <- rt7 %>% 
  select(-taxonomic_unit) %>% 
  # bring early
  relocate(include,
           geographic_scale:shift_source) %>% 
  # bring middle
  relocate(reference_no,
           publication_id:data_section,
           .after = language) %>% 
  # bring later
  relocate(comment, edit_info, recorder_name,
           .after = last_col()) %>% 
  arrange(publication_id,
          gradient, 
          position,
          kingdom,
          phylum,
          class,
          order,
          family,
          genus)


# ------------ #
# check any issues in key columns

# glimpse(rt_reorg)

keycols <- c("taxonomic_unit_revised",
             "kingdom",
             "ecosystem",
             "realm",
             "gradient",
             "position",
             "observation",
             "startperiod_startyear",
             "startperiod_midyear",
             "endperiod_endyear",
             "endperiod_midyear",
             "n_years",
             "number_of_censuses",
             "magnitude_of_shift",
             "unit_of_shift",
             "language",
             "publication_id")


# display cols that have na
rt_reorg %>%
  summarise(
    across(
      all_of(keycols),
      ~ sum(is.na(.x))
    )
  ) %>%
  pivot_longer(
    everything(),
    names_to = "column",
    values_to = "n_na"
  ) %>%
  filter(n_na > 0)



# ------------ #
rt_cleaned <- rt_reorg





# ------------ #
# save

readr::write_csv(
  rt_cleaned,
  fs::path(dir_out, "rt_cleaned.csv")
)

```



# Check cleaned data

```{r}
# ============================================== #
# check: show unique values for each column

nrow(rt_cleaned)

for (nm in names(rt_cleaned)) {
  vals <- unique(rt_cleaned[[nm]])
  
  cat("\n====================\n")
  cat("Column:", nm, "\n")
  cat("n_unique:", length(vals), "\n")
  cat("====================\n")
  
  print(vals)
  
}


# check entries for shift_signif when shift_tested == yes
rt_cleaned %>%
  filter(shift_tested == "yes") %>%
  pull(shift_significant) %>%
  unique()



```



# Visualize

## Tabulation

```{r}
# ============================================== #
# tabulation

rt_temp <- rt_cleaned %>%
  
  group_by(
    publication_id,
    language,
    realm,
    ecosystem,
    information_source,
    observation
  ) %>%
  
  summarise(
    
    study_area = str_c(
      sort(unique(study_area)),
      collapse = ", "
    ),
    
    specific_area = str_c(
      sort(unique(specific_area)),
      collapse = ", "
    ),
    
    area = str_c(specific_area, ", ", study_area),
    
    focus = str_c(
      sort(unique(kingdom)),
      collapse = ", "
    ),
    
    n_taxa = n_distinct(taxonomic_unit_revised),
    
    gradient = str_c(
      sort(unique(gradient)),
      collapse = ", "
    ),
    
    position = str_c(
      sort(unique(position)),
      collapse = ", "
    ),
    
    .groups = "drop"
  ) %>%
  
  arrange(
    realm,
    ecosystem,
    language
  ) %>% 
  
  relocate(study_area, specific_area, area,
           .after = language) %>% 
  
  relocate(focus, observation, n_taxa, information_source,
           .after = ecosystem) %>% 
  
  select(-specific_area, -study_area)





# ============================================== #
# Prepare table

# rt_temp$publication_id
# rt_temp$area


rt_temp_tab <- rt_temp %>%
  
  mutate(
    
    
    # ---------------------------------------------- #
    # area
    
    area = recode(
      area,
      "Mt.Fuji, Shizuoka" = "Mt. Fuji, Shizuoka",
      "Hokkaido, Hokkaido" = "Central Hokkaido",
      "Daisetsuzan National Park, Hokkaido" = "Daisetsuzan, Hokkaido",
      "Hakkoda Mountains, Aomori" = "Hakkoda Mountains, Aomori",
      "Mt.Rishiri, Hokkaido" = "Mt. Rishiri, Hokkaido",
      "Goyozan mountain, Iwate" = "Goyozan Mountains, Iwate",
      "Japan, Japan" = "Japan",
      "Kii, Kii" = "Kii Region",
      "West Japan, West Japan" = "West Japan",
      "Pacific coast, Tohoku" = "Tohoku Region",
      "Korea Strait, Japan" = "Korea Strait",
      "Miyagi, Tohoku" = "Miyagi Prefecture",
      "central-western sector, Sea of Japan" = "Sea of Japan",
      "Eastern coast, Japan, Western coast, Japan" = "East and West of Japan",
      "East China, Pacific coast, East Asia, Japan" = "East and West of Japan"
    ),
    
    
    
    # ---------------------------------------------- #
    # abbreviations
    
    realm = str_replace_all(
      realm,
      c(
        "terrestrial" = "t",
        "marine" = "m"
      )
    ),
    
    
    observation = str_replace_all(
      observation,
      c(
        "abundance" = "abu",
        "occurrence" = "occ"
      )
    ),
    
    focus = str_replace_all(
      focus,
      c(
        "Plantae" = "Pla",
        "Animalia" = "Ani",
        "Chromista" = "Chr"
      )
    ),
    
    gradient = str_replace_all(
      gradient,
      c(
        "elevation" = "Ele",
        "latitude" = "Lat",
        "longitude" = "Lon",
        "horizontal" = "Hor"
      )
    ),
    
    position = str_replace_all(
      position,
      c(
        "leading edge" = "Lea",
        "trailing edge" = "Tra",
        "centroid" = "Cen"
      )
    ),
    
    
    # ---------------------------------------------- #
    # sentence case
    
    language = str_to_sentence(language),
    realm = factor(
      str_to_sentence(as.character(realm)),
      levels = c("T", "M")
    ),
    ecosystem = str_to_sentence(as.character(ecosystem)),
    information_source = str_to_sentence(information_source),
    observation = str_to_sentence(observation)
  ) %>%
  
  arrange(
    realm,
    ecosystem,
    language
  )  %>% 
  
  relocate(realm) %>%
  
  # ---------------------------------------------- #
  # readable column names
  
  rename(
    "Publication" = publication_id,
    "Language" = language,
    "Study area" = area,
    "Realm" = realm,
    "Ecosystem" = ecosystem,
    "Information source" = information_source,
    "Observation" = observation,
    "Focus" = focus,
    "No. taxa" = n_taxa,
    "Gradient" = gradient,
    "Position" = position
  )

# ============================================== #
# Table

rt_temp_gt <- rt_temp_tab %>%
  gt() %>%
  
  cols_align(
    align = "left",
    columns = everything()
  ) %>%
  
  # cols_align(
  #   align = "center",
  #   columns = c(
  #     Language,
  #     Focus,
  #     `No. taxa`,
  #     Gradient,
  #     Position
  #   )
  # ) %>%
  
  tab_options(
    table.font.size = 12,
    data_row.padding = px(4)
  )

rt_temp_gt

gtsave(
  rt_temp_gt,
  filename = file.path(dir_out, "rt_temp_gt.png"),
  zoom = 2
)
```




## Count studies

### for prisma 

```{r}
# ============================================== #
# count studies

rt_temp <- rt_cleaned %>%
  distinct(realm, language, publication_id) %>% 
  arrange(realm, language, publication_id)

rt_temp %>%
  count(realm, language, name = "n_studies")




```


## Count observations

### rt_countobs

```{r}

rt_countobs <- rt_cleaned %>%
  mutate(
    across(
      c(
        realm,
        kingdom,
        class,
        gradient,
        position,
        information_source
      ),
      stringr::str_to_sentence
    ),
    
    realm = factor(
      realm,
      levels = c(
        "Terrestrial",
        "Marine"
      )
    ),
    
    gradient = factor(
      gradient,
      levels = c(
        "Elevation",
        "Latitude",
        "Longitude",
        "Horizontal"
      )
    )
  )


rt_countobs2 <- rt_cleaned %>%
  mutate(
    across(
      c(
        realm,
        kingdom,
        class,
        gradient,
        position,
        information_source
      ),
      stringr::str_to_sentence
    ),
    realm = factor(
      realm,
      levels = c(
        "Marine",
        "Terrestrial"
      )
    )
  )
```



### color coding

```{r}

# ============================================== #
# Realm color palettes

realm_palette <- function(realm, n) {
  
  # generate 1.5x as many colors
  # generate more colors when n = 1
  n_full <- if (n == 1) {
    ceiling(n * 3)
  } else {
    ceiling(n * 1.5)
  }
  
  
  # ---------------------------------------------- #
  # create full palette
  
  if (realm == "Marine") {
    
    cols <- grDevices::hcl.colors(
      n_full,
      palette = "Blues 3"
    )
    
  } else if (realm == "Terrestrial") {
    
    cols <- grDevices::hcl.colors(
      n_full,
      palette = "YlOrRd"
    )
    
  } else {
    
    cols <- grDevices::hcl.colors(
      n_full,
      palette = "Dark 3"
    )
  }
  
  
  # ---------------------------------------------- #
  # extract n colors around the middle
  
  start <- floor((n_full - n) / 2) + 1
  end <- start + n - 1
  
  cols[start:end]
}


# example
marine_colors <- realm_palette(
  "Marine",
  5
)

terrestrial_colors <- realm_palette(
  "Terrestrial",
  7
)

# scales::show_col(marine_colors)
# scales::show_col(terrestrial_colors)

```



### per realm

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# Realm colors

realm_colors <- c(
  "Marine" = realm_palette(
    "Marine",
    1
  ),
  "Terrestrial" = realm_palette(
    "Terrestrial",
    1
  )
)


# ============================================== #
# stacked bar plot: proportion per realm

rt_temp <- rt_countobs2 %>%
  count(realm) %>%
  mutate(
    prop = n / sum(n)
  )



# ============================================== #
# Plot
g_realm <- ggplot(
  rt_temp,
    aes(
      x = 1,
      y = prop,
      fill = realm
    )
  ) +
  
  geom_col(
    width = 0.7
  ) +
  
  coord_flip() +
  
  scale_fill_manual(
    values = realm_colors
  ) +
  
  scale_y_continuous(
    labels = scales::percent
  ) +
  
  labs(
    x = NULL,
    y = "Proportion",
    fill = "Realm"
  ) +
  
  theme_bw(base_size = 16) +
  
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# g_realm
```


#### refine 1: NOT USE


```{r}
#| fig-width: 8
#| fig-height: 2

g_realm <- ggplot(
  rt_temp,
  aes(
    x = 1,
    y = prop,
    fill = realm
  )
) +
  
  geom_col(
    width = 0.7
  ) +
  
  # Marine label inside the bar
  geom_text(
    data = rt_temp %>%
      filter(realm == "Marine"),
    aes(
      y = 1 - prop / 2,
      label = paste0(
        realm,
        " (",
        scales::percent(prop, accuracy = 1),
        ")"
      )
    ),
    size = 5
  ) +
  
  # connector from Terrestrial segment
  geom_segment(
    data = rt_temp %>%
      filter(realm == "Terrestrial"),
    aes(
      x = 1.2,
      xend = 1.45,
      y = prop / 2,
      yend = prop / 2 + 0.02
    ),
    inherit.aes = FALSE,
    linewidth = 0.6
  ) +
  
  # Terrestrial label above the bar and shifted right
  geom_text(
    data = rt_temp %>%
      filter(realm == "Terrestrial"),
    aes(
      x = 1.45,
      y = prop / 2 + 0.02,
      label = paste0(
        realm,
        " (",
        scales::percent(prop, accuracy = 1),
        ")"
      )
    ),
    inherit.aes = FALSE,
    size = 5,
    hjust = 0
  ) +
  
  coord_flip(
    clip = "off"
  ) +
  
  scale_fill_manual(
    values = realm_colors
  ) +
  
  theme_void() +
  
  theme(
    legend.position = "none",
    plot.margin = margin(
      t = 40,
      r = 10,
      b = 10,
      l = 10
    )
  )

g_realm

```



#### refine 2: USE


```{r}
#| fig-width: 8
#| fig-height: 2


# ============================================== #
# realm colors

realm_colors <- grDevices::hcl.colors(
  2,
  palette = "Pastel 1"
)

names(realm_colors) <- c(
  "Marine",
  "Terrestrial"
)


# ============================================== #
# Plot

g_realm <- ggplot(
  rt_temp,
  aes(
    x = 1,
    y = prop,
    fill = realm
  )
) +
  
  geom_col(
    width = 0.7,
    color = "black",
    linewidth = 1.0
  ) +
  
  # Marine label inside the bar
  geom_text(
    data = rt_temp %>%
      filter(realm == "Marine"),
    aes(
      y = 1 - prop / 2,
      label = paste0(
        realm,
        " (",
        scales::percent(prop, accuracy = 1),
        ")"
      )
    ),
    size = 7
  ) +
  
  # connector from Terrestrial segment
  geom_segment(
    data = rt_temp %>%
      filter(realm == "Terrestrial"),
    aes(
      x = 1.2,
      xend = 1.45,
      y = prop / 2,
      yend = prop / 2 + 0.02
    ),
    inherit.aes = FALSE,
    linewidth = 0.6
  ) +
  
  # Terrestrial label
  geom_text(
    data = rt_temp %>%
      filter(realm == "Terrestrial"),
    aes(
      x = 1.45,
      y = prop / 2 + 0.02,
      label = paste0(
        realm,
        " (",
        scales::percent(prop, accuracy = 1),
        ")"
      )
    ),
    inherit.aes = FALSE,
    size = 7,
    hjust = 0
  ) +
  
  ggtitle("a) Realm") +
  
  coord_flip(
    clip = "off"
  ) +
  
  scale_y_continuous(
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  
  scale_fill_manual(
    values = realm_colors
  ) +
  
  theme_void(
    base_size = 18
  ) +
  
  
  theme(
    legend.position = "none",
    
    plot.title = element_text(
      margin = margin(
        b = 10
      )
    ),
    plot.margin = margin(
      t = 10,
      r = 0,
      b = 0,
      l = 0
    )
  )

g_realm
```



### per taxonomic group

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# stacked bar: taxa composition

rt_temp <- rt_countobs %>%
  count(
    realm,
    kingdom,
    class
  )


bar_class <- rt_temp %>%
  group_by(realm) %>%
  arrange(kingdom, class, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


bar_kingdom_boundary <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    boundary = max(xmax),
    .groups = "drop"
  ) %>%
  filter(
    boundary < 1
  )




# ============================================== #
# Plot

g_taxo <- ggplot() +
  
  # ---------------------------------------------- #
  # class segments
  
  geom_rect(
    data = bar_class,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = class
    ),
    color = "white",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between kingdoms
  
  geom_segment(
    data = bar_kingdom_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    expand = c(0, 0)
  ) +
  
  labs(
    x = "Proportion",
    y = NULL,
    fill = "Class"
  ) +
  
  theme_bw(base_size = 16) +
  
  theme(
    panel.grid = element_blank(),
    strip.background = element_rect(
      fill = "grey90"
    )
  )

# g_taxo

```


#### refine color 1: NOT USE

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# prepare data

rt_temp <- rt_countobs %>%
  count(
    realm,
    kingdom,
    class
  )


bar_class <- rt_temp %>%
  group_by(realm) %>%
  arrange(kingdom, class, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0),
    
    # separate fill category by realm
    fill_group = paste(
      realm,
      class,
      sep = "__"
    )
  ) %>%
  ungroup()


# ============================================== #
# kingdom boundaries

bar_kingdom_boundary <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    boundary = max(xmax),
    .groups = "drop"
  ) %>%
  filter(
    boundary < 1
  )


# ============================================== #
# class colors by realm

class_colors_df <- bar_class %>%
  distinct(
    realm,
    class,
    fill_group
  ) %>%
  group_by(realm) %>%
  arrange(class, .by_group = TRUE) %>%
  mutate(
    color = case_when(
      
      realm == "Marine" ~
        realm_palette(
          "Marine",
          n()
        ),
      
      realm == "Terrestrial" ~
        realm_palette(
          "Terrestrial",
          n()
        )
    )
  ) %>%
  ungroup()


class_colors <- class_colors_df$color
names(class_colors) <- class_colors_df$fill_group





# ============================================== #
# Plot

g_taxo <- ggplot() +
  
  # ---------------------------------------------- #
  # class segments
  
  geom_rect(
    data = bar_class,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = fill_group
    ),
    color = "white",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between kingdoms
  
  geom_segment(
    data = bar_kingdom_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = class_colors,
    
    # remove realm prefix from legend labels
    labels = function(x) {
      stringr::str_remove(
        x,
        "^.*__"
      )
    }
  ) +
  
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    expand = c(0, 0)
  ) +
  
  labs(
    x = "Proportion",
    y = NULL,
    fill = "Class"
  ) +
  
  theme_void(base_size = 16) +
  
  theme(
    legend.position = "top",
    panel.grid = element_blank(),
    strip.background = element_rect(
      fill = "grey90"
    )
  )


g_taxo

```




#### refine color 2: NOT USE

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 8
#| fig-height: 2


# ============================================== #
# prepare data

rt_temp <- rt_countobs %>%
  count(
    realm,
    kingdom,
    class
  )


bar_class <- rt_temp %>%
  group_by(realm) %>%
  arrange(kingdom, class, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


# ============================================== #
# kingdom boundaries

bar_kingdom_boundary <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    boundary = max(xmax),
    .groups = "drop"
  ) %>%
  filter(
    boundary < 1
  )


# ============================================== #
# kingdom label positions

bar_kingdom_label <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# class colors

n_class <- bar_class %>%
  distinct(class) %>%
  nrow()

class_colors <- grDevices::hcl.colors(
  n_class,
  palette = "Pastel 1"
)

names(class_colors) <- sort(
  unique(bar_class$class)
)


# ============================================== #
# Plot

g_taxo <- ggplot() +
  
  # ---------------------------------------------- #
  # class segments
  
  geom_rect(
    data = bar_class,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = class
    ),
    color = "white",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between kingdoms
  
  geom_segment(
    data = bar_kingdom_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # kingdom labels
  
  geom_text(
    data = bar_kingdom_label,
    aes(
      x = x,
      y = 1.05,
      label = kingdom
    ),
    inherit.aes = FALSE,
    size = 5
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = class_colors
  ) +
  
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(0, 1.2),
    expand = c(0, 0)
  ) +
  
  labs(
    x = "Proportion",
    y = NULL,
    fill = "Class"
  ) +
  
  theme_void(base_size = 16) +
  
  theme(
    legend.position = "top",
    panel.grid = element_blank(),
    strip.text = element_text(
      size = 18,
      face = "bold"
    )
)


# g_taxo

```




#### refine 3: NOT USE

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# class level

bar_class <- rt_countobs %>%
  count(
    realm,
    kingdom,
    class
  ) %>%
  group_by(realm) %>%
  arrange(kingdom, class, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


# ============================================== #
# kingdom level

bar_kingdom <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# kingdom boundaries within class bar

bar_kingdom_boundary <- bar_kingdom %>%
  group_by(realm) %>%
  arrange(xmin, .by_group = TRUE) %>%
  filter(
    xmax < 1
  ) %>%
  transmute(
    realm,
    boundary = xmax
  ) %>%
  ungroup()


# ============================================== #
# class label positions

bar_class_label <- bar_class %>%
  mutate(
    x = (xmin + xmax) / 2
  )


# ============================================== #
# large classes: label inside bar

bar_class_label_large <- bar_class_label %>%
  filter(
    prop >= 0.1
  )


# ============================================== #
# small classes: labels outside bar

bar_class_label_small <- bar_class_label %>%
  filter(
    prop < 0.1
  ) %>%
  group_by(realm) %>%
  arrange(x, .by_group = TRUE) %>%
  mutate(
    label_id = row_number(),
    
    # alternate between two rows
    label_row = (label_id - 1) %% 2,
    
    # spread labels horizontally
    label_x = seq(
      0.08,
      0.92,
      length.out = n()
    ),
    
    # stagger vertically
    label_y = -0.18 - 0.18 * label_row
  ) %>%
  ungroup()


# ============================================== #
# class colors

n_class <- bar_class %>%
  distinct(class) %>%
  nrow()

class_colors <- grDevices::hcl.colors(
  n_class,
  palette = "Pastel 1"
)

names(class_colors) <- sort(
  unique(bar_class$class)
)


# ============================================== #
# Plot

g_taxo <- ggplot() +
  
  # ---------------------------------------------- #
  # kingdom bar
  
  geom_rect(
    data = bar_kingdom,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 1,
      ymax = 1.4
    ),
    fill = "white",
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # kingdom text inside kingdom bar
  
  geom_text(
    data = bar_kingdom,
    aes(
      x = x,
      y = 1.2,
      label = kingdom
    ),
    inherit.aes = FALSE,
    size = 4.5
  ) +
  
  # ---------------------------------------------- #
  # class bar
  
  geom_rect(
    data = bar_class,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = class
    ),
    color = "black",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between kingdoms
  # within class bar
  
  geom_segment(
    data = bar_kingdom_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # thicker outer border around entire class bar
  
  geom_rect(
    data = bar_class %>%
      distinct(realm),
    aes(
      xmin = 0,
      xmax = 1,
      ymin = 0,
      ymax = 1
    ),
    inherit.aes = FALSE,
    fill = NA,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # large class labels inside class bar
  
  geom_text(
    data = bar_class_label_large,
    aes(
      x = x,
      y = 0.5,
      label = class
    ),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  
  
  # ---------------------------------------------- #
  # diagonal connectors for small classes
  
  geom_segment(
    data = bar_class_label_small,
    aes(
      x = x,
      xend = label_x,
      y = 0,
      yend = label_y + 0.04
    ),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # small class labels outside class bar
  
  geom_text(
    data = bar_class_label_small,
    aes(
      x = label_x,
      y = label_y,
      label = class
    ),
    inherit.aes = FALSE,
    size = 3.5,
    hjust = 0.5
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = class_colors
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(-0.5, 1.4),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # labels
  
  ggtitle("b) Taxonomic group (kingdom and class)") +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  
  # ============================================== #
  # theme
  
  theme_void(
    base_size = 18
  ) +
  
  theme(
    legend.position = "none",
    
    plot.title = element_text(
      margin = margin(
        b = 10
      )
    ),
    
    strip.text = element_text(
      size = 18,
      face = "bold"
    ),
    plot.margin = margin(
      t = 10,
      r = 0,
      b = 0,
      l = 0
    )
  )


g_taxo

```




#### refine 4: USE

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# class level

bar_class <- rt_countobs %>%
  count(
    realm,
    kingdom,
    class
  ) %>%
  group_by(realm) %>%
  arrange(kingdom, class, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


# ============================================== #
# kingdom level

bar_kingdom <- bar_class %>%
  group_by(
    realm,
    kingdom
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# kingdom label positions

bar_kingdom_label <- bar_kingdom %>%
  mutate(
    prop = xmax - xmin
  )


# ============================================== #
# large kingdoms: label inside bar

bar_kingdom_label_large <- bar_kingdom_label %>%
  filter(
    prop >= 0.1
  )


# ============================================== #
# small kingdoms: label outside bar

bar_kingdom_label_small <- bar_kingdom_label %>%
  filter(
    prop < 0.1
  ) %>%
  group_by(realm) %>%
  arrange(x, .by_group = TRUE) %>%
  mutate(
    label_id = row_number(),
    label_row = (label_id - 1) %% 2,
    
    # move labels at edges toward the center
    label_x = case_when(
      x < 0.1 ~ x + 0.08,
      x > 0.9 ~ x - 0.08,
      TRUE ~ x
    ),
    
    label_y = 1.62 + 0.18 * label_row
  ) %>%
  ungroup()



# ============================================== #
# kingdom boundaries within class bar

bar_kingdom_boundary <- bar_kingdom %>%
  group_by(realm) %>%
  arrange(xmin, .by_group = TRUE) %>%
  filter(
    xmax < 1
  ) %>%
  transmute(
    realm,
    boundary = xmax
  ) %>%
  ungroup()


# ============================================== #
# class label positions

bar_class_label <- bar_class %>%
  mutate(
    x = (xmin + xmax) / 2
  )


# ============================================== #
# large classes: label inside bar

bar_class_label_large <- bar_class_label %>%
  filter(
    prop >= 0.1
  )


# ============================================== #
# small classes: labels outside bar

bar_class_label_small <- bar_class_label %>%
  filter(
    prop < 0.1
  ) %>%
  group_by(realm) %>%
  arrange(x, .by_group = TRUE) %>%
  mutate(
    label_id = row_number(),
    
    # alternate between two rows
    label_row = (label_id - 1) %% 2,
    
    # spread labels horizontally
    label_x = seq(
      0.08,
      0.92,
      length.out = n()
    ),
    
    # stagger vertically
    label_y = -0.18 - 0.18 * label_row
  ) %>%
  ungroup() %>% 
  mutate(
    # edit manually
    label_x = case_when(
      class == "Liliopsida" ~ 0.5,
      class == "Bivalvia" ~ 0.06,
      class == "Cephalopoda" ~ 0.14, 
      class == "Elasmobranchii" ~ 0.24, 
      class == "Holothuroidea" ~ 0.34, 
      class == "Malacostraca" ~ 0.54, 
      class == "Polyplacophora" ~ 0.6, 
      class == "Florideophyceae" ~ 0.86, 
      
      TRUE ~ label_x
    ),
    
    # edit manually
    label_y = case_when(
      class == "Holothuroidea" ~ -0.36, 
      class == "Malacostraca" ~ -0.36, 
      class == "Polyplacophora" ~ -0.18, 
      
      TRUE ~ label_y
    )
  )


# ============================================== #
# class colors

n_class <- bar_class %>%
  distinct(class) %>%
  nrow()

class_colors <- grDevices::hcl.colors(
  n_class,
  palette = "Pastel 1"
)

names(class_colors) <- sort(
  unique(bar_class$class)
)


# ============================================== #
# Plot

g_taxo <- ggplot() +
  
  # ---------------------------------------------- #
  # kingdom bar
  
  geom_rect(
    data = bar_kingdom,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 1,
      ymax = 1.4
    ),
    fill = "white",
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # large kingdom labels inside kingdom bar
  
  geom_text(
    data = bar_kingdom_label_large,
    aes(
      x = x,
      y = 1.2,
      label = kingdom
    ),
    inherit.aes = FALSE,
    size = 6
  ) +
  
  
  # ---------------------------------------------- #
  # connectors for small kingdoms
  
  geom_segment(
    data = bar_kingdom_label_small,
    aes(
      x = x,
      xend = label_x,
      y = 1.4,
      yend = label_y - 0.1
    ),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # small kingdom labels outside kingdom bar
  
  geom_text(
    data = bar_kingdom_label_small,
    aes(
      x = label_x,
      y = label_y,
      label = kingdom
    ),
    inherit.aes = FALSE,
    size = 6,
    hjust = 0.5
  ) +
  
  # ---------------------------------------------- #
  # class bar
  
  geom_rect(
    data = bar_class,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = class
    ),
    color = "black",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between kingdoms
  # within class bar
  
  geom_segment(
    data = bar_kingdom_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # thicker outer border around entire class bar
  
  geom_rect(
    data = bar_class %>%
      distinct(realm),
    aes(
      xmin = 0,
      xmax = 1,
      ymin = 0,
      ymax = 1
    ),
    inherit.aes = FALSE,
    fill = NA,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # large class labels inside class bar
  
  geom_text(
    data = bar_class_label_large,
    aes(
      x = x,
      y = 0.5,
      label = class
    ),
    inherit.aes = FALSE,
    size = 4.5
  ) +
  
  
  # ---------------------------------------------- #
  # diagonal connectors for small classes
  
  geom_segment(
    data = bar_class_label_small,
    aes(
      x = x,
      xend = label_x,
      y = 0,
      yend = label_y + 0.06
    ),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # small class labels outside class bar
  
  geom_text(
    data = bar_class_label_small,
    aes(
      x = label_x,
      y = label_y,
      label = class
    ),
    inherit.aes = FALSE,
    size = 4.5,
    hjust = 0.5
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = class_colors
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(-0.5, 1.75),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # labels
  
  ggtitle("b) Taxonomic group (kingdom and class)") +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  
  # ============================================== #
  # theme
  
  theme_void(
    base_size = 18
  ) +
  
  theme(
    legend.position = "none",
    
    plot.title = element_text(
      margin = margin(
        b = 10
      )
    ),
    
    strip.text = element_text(
      size = 18,
      face = "bold"
    )
  )


g_taxo

```




### per gradient x position

#### orig: NOT USE

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# stacked bar: gradient > position

bar_position <- rt_countobs %>%
  
  count(
    realm,
    gradient,
    position
  ) %>%
  
  group_by(realm) %>%
  arrange(gradient, position, .by_group = TRUE) %>%
  
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  
  ungroup()

# ============================================== #
# gradient boundaries

bar_gradient_boundary <- bar_position %>%
  group_by(
    realm,
    gradient
  ) %>%
  summarise(
    boundary = max(xmax),
    .groups = "drop"
  ) %>%
  filter(
    boundary < 1
  )

# ============================================== #
# gradient label positions

bar_gradient_label <- bar_position %>%
  group_by(
    realm,
    gradient
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# position colors

n_position <- n_distinct(
  bar_position$position
)

position_colors <- grDevices::hcl.colors(
  n_position,
  palette = "Pastel 1"
)

scales::show_col(position_colors)

names(position_colors) <- sort(
  unique(bar_position$position)
)


# ============================================== #
# Plot

g_grad <- ggplot() +
  
  # ---------------------------------------------- #
  # small boundaries: position
  
  geom_rect(
    data = bar_position,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = position
    ),
    color = "white",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # large boundaries: gradient
  
  geom_segment(
    data = bar_gradient_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # gradient labels
  
  geom_text(
    data = bar_gradient_label,
    aes(
      x = x,
      y = 1.05,
      label = gradient
    ),
    inherit.aes = FALSE,
    size = 5
  ) +
  
  
  # ============================================== #
  # realm
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = position_colors
  ) +
  
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(0, 1.2),
    expand = c(0, 0)
  ) +
  
  
  labs(
    x = "Proportion",
    y = NULL,
    fill = "Position"
  ) +
  
  theme_void(base_size = 16) +
  
  theme(
    legend.position = "top",
    panel.grid = element_blank(),
    strip.text = element_text(
      size = 18,
      face = "bold"
    )
  )


# g_grad
```


#### refine 1: NOT USE

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# position level

bar_position <- rt_countobs %>%
  count(
    realm,
    gradient,
    position
  ) %>%
  group_by(realm) %>%
  arrange(gradient, position, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


# ============================================== #
# gradient level

bar_gradient <- bar_position %>%
  group_by(
    realm,
    gradient
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# gradient boundaries within position bar

bar_gradient_boundary <- bar_gradient %>%
  group_by(realm) %>%
  arrange(xmin, .by_group = TRUE) %>%
  filter(
    xmax < 1
  ) %>%
  transmute(
    realm,
    boundary = xmax
  ) %>%
  ungroup()


# ============================================== #
# position colors

n_position <- n_distinct(
  bar_position$position
)

position_colors <- grDevices::hcl.colors(
  n_position,
  palette = "Pastel 1"
)

names(position_colors) <- sort(
  unique(bar_position$position)
)


# ============================================== #
# Plot

g_grad <- ggplot() +
  
  # ---------------------------------------------- #
  # gradient bar
  
  geom_rect(
    data = bar_gradient,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 1,
      ymax = 1.4
    ),
    fill = "white",
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # gradient text inside gradient bar
  
  geom_text(
    data = bar_gradient,
    aes(
      x = x,
      y = 1.2,
      label = gradient
    ),
    inherit.aes = FALSE,
    size = 4.5
  ) +
  
  
  # ---------------------------------------------- #
  # position bar
  
  geom_rect(
    data = bar_position,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = position
    ),
    color = "black",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between gradients
  # within position bar
  
  geom_segment(
    data = bar_gradient_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # thicker outer border around entire position bar
  
  geom_rect(
    data = bar_position %>%
      distinct(realm),
    aes(
      xmin = 0,
      xmax = 1,
      ymin = 0,
      ymax = 1
    ),
    inherit.aes = FALSE,
    fill = NA,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = position_colors
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(0, 1.4),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # labels
  
  ggtitle("c) Gradient and position") +
  
  labs(
    x = NULL,
    y = NULL,
    fill = "Position"
  ) +
  
  
  # ============================================== #
  # theme
  
  theme_void(
    base_size = 16
  ) +
  
  theme(
    legend.position = "bottom",
    
    
    plot.title = element_text(
      margin = margin(
        b = 10
      )
    ),
    
    strip.text = element_text(
      size = 18,
      face = "bold"
    ),
    plot.margin = margin(
      t = 10,
      r = 0,
      b = 0,
      l = 0
    )
  )


g_grad
```



#### refine 2: USE

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# position level

bar_position <- rt_countobs %>%
  count(
    realm,
    gradient,
    position
  ) %>%
  group_by(realm) %>%
  arrange(gradient, position, .by_group = TRUE) %>%
  mutate(
    prop = n / sum(n),
    xmax = cumsum(prop),
    xmin = lag(xmax, default = 0)
  ) %>%
  ungroup()


# ============================================== #
# gradient level

bar_gradient <- bar_position %>%
  group_by(
    realm,
    gradient
  ) %>%
  summarise(
    xmin = min(xmin),
    xmax = max(xmax),
    x = (xmin + xmax) / 2,
    .groups = "drop"
  )


# ============================================== #
# gradient label positions

bar_gradient_label <- bar_gradient %>%
  mutate(
    prop = xmax - xmin
  )


# ============================================== #
# large gradients: label inside bar

bar_gradient_label_large <- bar_gradient_label %>%
  filter(
    prop >= 0.15
  )


# ============================================== #
# small gradients: label outside bar

bar_gradient_label_small <- bar_gradient_label %>%
  filter(
    prop < 0.15
  ) %>%
  group_by(realm) %>%
  arrange(x, .by_group = TRUE) %>%
  mutate(
    label_id = row_number(),
    label_row = (label_id - 1) %% 2,
    
    # shift edge labels slightly inward
    label_x = case_when(
      x < 0.1 ~ x + 0.08,
      x > 0.9 ~ x - 0.08,
      TRUE ~ x
    ),
    
    label_y = 1.62 + 0.18 * label_row
  ) %>%
  ungroup() %>% 
  mutate(
    # edit manually
    label_x = case_when(
      realm == "Terrestrial" & gradient == "Longitude" ~ 0.84,
      realm == "Terrestrial" & gradient == "Horizontal" ~ 0.9,
      TRUE ~ label_x
    )
  )
  

# ============================================== #
# gradient boundaries within position bar

bar_gradient_boundary <- bar_gradient %>%
  group_by(realm) %>%
  arrange(xmin, .by_group = TRUE) %>%
  filter(
    xmax < 1
  ) %>%
  transmute(
    realm,
    boundary = xmax
  ) %>%
  ungroup()


# ============================================== #
# position colors

n_position <- n_distinct(
  bar_position$position
)

position_colors <- grDevices::hcl.colors(
  n_position,
  palette = "Pastel 1"
)

names(position_colors) <- sort(
  unique(bar_position$position)
)


# ============================================== #
# Plot

g_grad <- ggplot() +
  
  # ---------------------------------------------- #
  # gradient bar
  
  geom_rect(
    data = bar_gradient,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 1,
      ymax = 1.4
    ),
    fill = "white",
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # large gradient labels inside gradient bar
  
  geom_text(
    data = bar_gradient_label_large,
    aes(
      x = x,
      y = 1.2,
      label = gradient
    ),
    inherit.aes = FALSE,
    size = 6
  ) +
  
  
  # ---------------------------------------------- #
  # connectors for small gradients
  
  geom_segment(
    data = bar_gradient_label_small,
    aes(
      x = x,
      xend = label_x,
      y = 1.4,
      yend = label_y - 0.1
    ),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # small gradient labels outside gradient bar
  
  geom_text(
    data = bar_gradient_label_small,
    aes(
      x = label_x,
      y = label_y,
      label = gradient
    ),
    inherit.aes = FALSE,
    size = 6,
    hjust = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # position bar
  
  geom_rect(
    data = bar_position,
    aes(
      xmin = xmin,
      xmax = xmax,
      ymin = 0,
      ymax = 1,
      fill = position
    ),
    color = "black",
    linewidth = 0.5
  ) +
  
  
  # ---------------------------------------------- #
  # thicker boundaries between gradients
  # within position bar
  
  geom_segment(
    data = bar_gradient_boundary,
    aes(
      x = boundary,
      xend = boundary,
      y = 0,
      yend = 1
    ),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ---------------------------------------------- #
  # thicker outer border around entire position bar
  
  geom_rect(
    data = bar_position %>%
      distinct(realm),
    aes(
      xmin = 0,
      xmax = 1,
      ymin = 0,
      ymax = 1
    ),
    inherit.aes = FALSE,
    fill = NA,
    color = "black",
    linewidth = 1.3
  ) +
  
  
  # ============================================== #
  # facets
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  
  # ============================================== #
  # colors
  
  scale_fill_manual(
    values = position_colors
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    breaks = NULL,
    limits = c(0, 1.9),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # labels
  
  ggtitle("c) Gradient and position") +
  
  labs(
    x = NULL,
    y = NULL,
    fill = "Position"
  ) +
  
  
  # ============================================== #
  # theme
  
  theme_void(
    base_size = 20
  ) +
  
  theme(
    legend.position = "bottom",
    
    
    plot.title = element_text(
      margin = margin(
        b = 10
      )
    ),
    
    strip.text = element_text(
      size = 18,
      face = "bold"
    )
  )


g_grad
```




### per information source

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# information source per realm

rt_temp <- rt_countobs %>%
  
  count(
    realm,
    information_source
  ) %>%
  
  group_by(realm) %>%
  
  mutate(
    prop = n / sum(n)
  ) %>%
  
  ungroup()


# ============================================== #
# Plot



g_info <- ggplot(
  rt_temp, 
  aes(
    x = 1,
    y = prop,
    fill = information_source
  )
) +
  
  geom_col(
    width = 0.7,
    color = "white",
    linewidth = 0.5
  ) +
  
  coord_flip() +
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  labs(
    x = NULL,
    y = "Proportion",
    fill = "Information source"
  ) +
  
  theme_bw(base_size = 16) +
  
  theme(
    legend.position = "top",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    strip.background = element_rect(
      fill = "grey90"
    )
  )



```


#### refine color: USE

```{r}
#| fig-width: 8
#| fig-height: 2

# ============================================== #
# information source colors

information_sources <- sort(
  unique(rt_countobs$information_source)
)

information_source_colors <- grDevices::hcl.colors(
  length(information_sources),
  palette = "Pastel 1"
)

names(information_source_colors) <- information_sources


# ============================================== #
# information source per realm

rt_temp <- rt_countobs %>%
  
  count(
    realm,
    information_source
  ) %>%
  
  group_by(realm) %>%
  
  mutate(
    prop = n / sum(n)
  ) %>%
  
  ungroup()

# ============================================== #
# Plot

g_info <- ggplot(
  rt_temp,
  aes(
    x = 1,
    y = prop,
    fill = information_source
  )
) +
  
  geom_col(
    width = 0.7,
    color = "black",
    linewidth = 0.7
  ) +
  
  coord_flip() +
  
  facet_wrap(
    ~ realm,
    ncol = 2
  ) +
  
  scale_fill_manual(
    values = information_source_colors
  ) +
  
  scale_y_continuous(
    labels = scales::percent,
    expand = c(0, 0)
  ) +
  
  ggtitle("d) Information source") +
  
  labs(
    x = NULL,
    y = NULL,
    fill = ""
  ) +
  
  theme_classic(
    base_size = 20
  ) +
  
  theme(
    legend.position = "bottom",
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    
    strip.background = element_blank(),
    strip.text = element_text(
      size = 18,
      face = "bold"
    )
  )

g_info
```



### Combine

```{r}
#| fig-width: 14
#| fig-height: 14

# g_realm
# g_taxo
# g_grad
# g_info

# ============================================== #
# Combine plots vertically

g_combined <- g_realm /
  plot_spacer() /
  g_taxo /
  plot_spacer() /
  g_grad /
  plot_spacer() /
  g_info +
  
  plot_layout(
    ncol = 1,
    heights = c(
      2,
      0.2,
      3.4,
      0.2,
      3,
      0.2,
      2
    )
  )

g_combined

```



## Time plot

### 1: row height varies with sample size

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 10
#| fig-height: 8


# ============================================== #
# Gantt chart-like plot of study duration per observation

rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_midyear),
    !is.na(endperiod_midyear)
  ) %>%
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(startperiod_midyear, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_midyear,
    endperiod_midyear
  ) %>%
  
  # publication legend follows arranged order
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    ),
    
    # unique observation ID across whole dataset
    obs_id = row_number()
  )%>%
  mutate(
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


class(rt_temp$obs_id)


ggplot(rt_temp, aes(y = obs_id, color = publication_id)) +
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = endperiod_midyear,
      yend = obs_id,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = startperiod_midyear+1,
      yend = obs_id
    ),
    color = "black",
    linewidth = 1,
    alpha = 1
  ) +
  geom_segment(
    aes(
      x = endperiod_midyear-1,
      xend = endperiod_midyear,
      yend = obs_id
    ),
    color = "black",
    linewidth = 1,
    alpha = 1
  ) +
  # geom_errorbar(
  #   aes(
  #     y = obs_id,
  #     xmin = startperiod_midyear,
  #     xmax = endperiod_midyear,
  #     linetype = census_linetype
  #   ),
  #   width = 0.3,
  #   linewidth = 1,
  #   orientation = "y",
  #   alpha = 1
  # ) +
  geom_text(
    aes(
      x = endperiod_midyear,
      label = number_of_censuses
    ),
    hjust = -0.3,
    color = "black",
    size = 3.5
  ) +
  scale_linetype_manual(
    values = c(
      "2 censuses" = "dotted",
      ">2 censuses" = "solid"
    )
  ) +
  # geom_point(
  #   aes(x = startperiod_midyear),
  #   size = 2,
  #   alpha = 1
  # ) +
  # geom_point(
  #   aes(x = endperiod_midyear),
  #   size = 2,
  #   alpha = 1
  # ) +
  
  facet_grid2(
    rows = vars(realm),
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  
  scale_x_continuous(
    expand = expansion(mult = c(0.02, 0.1))
  ) +
  
  scale_y_reverse() +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  labs(
    x = "Year",
    y = "Observation",
    color = "Reference no",
    title = "Years between the original survey and resurvey"
  ) +
  
  guides(
    color = guide_legend(ncol = 1)
  ) +
  
  theme_bw(base_size = 16) +
  theme(
    legend.position = "right",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(hjust = 0),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )


```


### 2: row height is fixed regardless of sample size

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 10
#| fig-height: 8

# ============================================== #
# Gantt chart-like plot of study duration per observation
#
# Equal vertical height for:
#   1. each realm
#   2. each reference within each realm


rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_midyear),
    !is.na(endperiod_midyear)
  ) %>%
  
  
  # ---------------------------------------------- #
  # First year represented by each publication
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(
      startperiod_midyear,
      na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Arrange publications and observations
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_midyear,
    endperiod_midyear
  ) %>%
  
  
  # ---------------------------------------------- #
  # Publication factor follows arranged order
  
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  ) %>%
  
  
  # ============================================== #
  # Give each reference an equal-height band
  # within each realm
  
  group_by(realm) %>%
  mutate(
    
    # reference order within realm
    reference_order = match(
      publication_id,
      unique(publication_id)
    ),
    
    # number of references in realm
    n_ref_realm = n_distinct(publication_id)
    
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Give each observation a position within
  # its reference band
  
  group_by(realm, publication_id) %>%
  mutate(
    
    n_entry = n(),
    obs_order = row_number(),
    
    # center of this reference's band
    reference_center = (
      reference_order - 0.5
    ) / n_ref_realm,
    
    # total vertical height available
    # to this reference
    reference_height = 1 / n_ref_realm,
    
    # observation positions within reference band
    #
    # observations occupy 70% of the available
    # reference band (±35% around the center)
    obs_offset = if (n() == 1) {
      
      0
      
    } else {
      
      (
        (obs_order - 1) / (n_entry - 1) - 0.5
      ) * 0.7 * first(reference_height)
      
    },
    
    # final y position
    obs_y = reference_center + obs_offset
    
  ) %>%
  
  ungroup() %>%
  
  mutate(
    realm = factor(
      realm,
      levels = c("terrestrial", "marine"),
      labels = c("Terrestrial", "Marine")
    ),
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


# ============================================== #
# Reference labels
#
# Because each reference now has its own
# equal-height band, simply use reference_center
# as the label position.


publication_labels <- rt_temp %>%
  
  group_by(
    realm,
    publication_id
  ) %>%
  
  summarise(
    label_y = first(reference_center),
    n_entry = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    publication_label = as.character(publication_id),
    
    publication_label = case_when(
      
      # References containing "et al."
      stringr::str_detect(publication_label, " et al\\. ") ~
        stringr::str_replace(
          publication_label,
          "^(.*?) et al\\. (.*)$",
          "'\\1 '*italic('et al.')*' \\2'"
        ),
      
      # References without "et al."
      TRUE ~ paste0(
        "'",
        publication_label,
        "'"
      )
    ),
    
    # Add italic (n = x)
    publication_label = paste0(
      publication_label,
      "*' ('*italic(n == ",
      n_entry,
      ")*')'"
    )
  )


# ============================================== #
# Publication colors


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


# ============================================== #
# Plot


ggplot(
  rt_temp,
  aes(
    y = obs_y,
    color = publication_id
  )
) +
  
  # ---------------------------------------------- #
  # Study period
  
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = endperiod_midyear,
      yend = obs_y,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at first survey year
  
  geom_text(
    aes(
      x = startperiod_midyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at final survey year
  
  geom_text(
    aes(
      x = endperiod_midyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Reference labels
  
  geom_text(
    data = publication_labels,
    aes(
      x = -Inf,
      y = label_y,
      label = publication_label
    ),
    inherit.aes = FALSE,
    parse = TRUE,
    hjust = 1.05,
    size = 3.5
  ) +
  
  
  # ============================================== #
  # Scales
  
  scale_linetype_manual(
    values = c(
      "2 censuses" = "longdash",
      ">2 censuses" = "solid"
    )
  ) +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  scale_x_continuous(
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  
  scale_y_reverse(
    limits = c(1.02, -0.02),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # Equal-height realm facets
  
  facet_grid2(
    rows = vars(realm),
    scales = "fixed",
    space = "fixed"
  ) +
  
  
  coord_cartesian(
    clip = "off"
  ) +
  
  
  labs(
    x = "Year of record",
    y = NULL
  ) +
  
  
  guides(
    color = "none",
    linetype = "none"
  ) +
  
  
  theme_bw(base_size = 16) +
  
  theme(
    legend.position = "none",
    
    strip.placement = "outside",
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    # remove y-axis grid lines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    panel.spacing.y = unit(0.15, "lines"),
    
    plot.margin = margin(
      t = 5.5,
      r = 5.5,
      b = 5.5,
      l = 150
    )
  )



```


### 3: row height is fixed regardless of sample size, start-start to end-end

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 10
#| fig-height: 8

# ============================================== #
# Gantt chart-like plot of study duration per observation
#
# Equal vertical height for:
#   1. each realm
#   2. each reference within each realm


rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_startyear),
    !is.na(endperiod_endyear)
  ) %>%
  
  
  # ---------------------------------------------- #
  # First year represented by each publication
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(
      startperiod_startyear,
      na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Arrange publications and observations
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_startyear,
    endperiod_endyear
  ) %>%
  
  
  # ---------------------------------------------- #
  # Publication factor follows arranged order
  
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  ) %>%
  
  
  # ============================================== #
  # Give each reference an equal-height band
  # within each realm
  
  group_by(realm) %>%
  mutate(
    
    # reference order within realm
    reference_order = match(
      publication_id,
      unique(publication_id)
    ),
    
    # number of references in realm
    n_ref_realm = n_distinct(publication_id)
    
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Give each observation a position within
  # its reference band
  
  group_by(realm, publication_id) %>%
  mutate(
    
    n_entry = n(),
    obs_order = row_number(),
    
    # center of this reference's band
    reference_center = (
      reference_order - 0.5
    ) / n_ref_realm,
    
    # total vertical height available
    # to this reference
    reference_height = 1 / n_ref_realm,
    
    # observation positions within reference band
    #
    # observations occupy 70% of the available
    # reference band (±35% around the center)
    obs_offset = if (n() == 1) {
      
      0
      
    } else {
      
      (
        (obs_order - 1) / (n_entry - 1) - 0.5
      ) * 0.7 * first(reference_height)
      
    },
    
    # final y position
    obs_y = reference_center + obs_offset
    
  ) %>%
  
  ungroup() %>%
  
  mutate(
    realm = factor(
      realm,
      levels = c("terrestrial", "marine"),
      labels = c("Terrestrial", "Marine")
    ),
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


# ============================================== #
# Reference labels
#
# Because each reference now has its own
# equal-height band, simply use reference_center
# as the label position.


publication_labels <- rt_temp %>%
  
  group_by(
    realm,
    publication_id
  ) %>%
  
  summarise(
    label_y = first(reference_center),
    n_entry = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    publication_label = as.character(publication_id),
    
    publication_label = case_when(
      
      # References containing "et al."
      stringr::str_detect(publication_label, " et al\\. ") ~
        stringr::str_replace(
          publication_label,
          "^(.*?) et al\\. (.*)$",
          "'\\1 '*italic('et al.')*' \\2'"
        ),
      
      # References without "et al."
      TRUE ~ paste0(
        "'",
        publication_label,
        "'"
      )
    ),
    
    # Add italic (n = x)
    publication_label = paste0(
      publication_label,
      "*' ('*italic(n == ",
      n_entry,
      ")*')'"
    )
  )


# ============================================== #
# Publication colors


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


# ============================================== #
# Plot


ggplot(
  rt_temp,
  aes(
    y = obs_y,
    color = publication_id
  )
) +
  
  # ---------------------------------------------- #
  # Study period
  
  geom_segment(
    aes(
      x = startperiod_startyear,
      xend = endperiod_endyear,
      yend = obs_y,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at first survey year
  
  geom_text(
    aes(
      x = startperiod_startyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at final survey year
  
  geom_text(
    aes(
      x = endperiod_endyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Reference labels
  
  geom_text(
    data = publication_labels,
    aes(
      x = -Inf,
      y = label_y,
      label = publication_label
    ),
    inherit.aes = FALSE,
    parse = TRUE,
    hjust = 1.05,
    size = 3.5
  ) +
  
  
  # ============================================== #
  # Scales
  
  scale_linetype_manual(
    values = c(
      "2 censuses" = "longdash",
      ">2 censuses" = "solid"
    )
  ) +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  scale_x_continuous(
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  
  scale_y_reverse(
    limits = c(1.02, -0.02),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # Equal-height realm facets
  
  facet_grid2(
    rows = vars(realm),
    scales = "fixed",
    space = "fixed"
  ) +
  
  
  coord_cartesian(
    clip = "off"
  ) +
  
  
  labs(
    x = "Year of record",
    y = NULL
  ) +
  
  
  guides(
    color = "none",
    linetype = "none"
  ) +
  
  
  theme_bw(base_size = 16) +
  
  theme(
    legend.position = "none",
    
    strip.placement = "outside",
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    # remove y-axis grid lines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    panel.spacing.y = unit(0.15, "lines"),
    
    plot.margin = margin(
      t = 5.5,
      r = 5.5,
      b = 5.5,
      l = 150
    )
  )



```

### 4: row height is fixed regardless of sample size, start-start to end-end, different color setting

```{r}
#| fig-width: 15
#| fig-height: 9

# ============================================== #
# Gantt chart-like plot of study duration per observation
#
# Equal vertical height for:
#   1. each realm
#   2. each reference within each realm


rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_startyear),
    !is.na(endperiod_endyear)
  ) %>%
  
  
  # ---------------------------------------------- #
  # First year represented by each publication
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(
      startperiod_startyear,
      na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Arrange publications and observations
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_startyear,
    endperiod_endyear
  ) %>%
  
  
  # ---------------------------------------------- #
  # Publication factor follows arranged order
  
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  ) %>%
  
  
  # ============================================== #
  # Give each reference an equal-height band
  # within each realm
  
  group_by(realm) %>%
  mutate(
    
    # reference order within realm
    reference_order = match(
      publication_id,
      unique(publication_id)
    ),
    
    # number of references in realm
    n_ref_realm = n_distinct(publication_id)
    
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Give each observation a position within
  # its reference band
  
  group_by(realm, publication_id) %>%
  mutate(
    
    n_entry = n(),
    obs_order = row_number(),
    
    # center of this reference's band
    reference_center = (
      reference_order - 0.5
    ) / n_ref_realm,
    
    # total vertical height available
    # to this reference
    reference_height = 1 / n_ref_realm,
    
    # observation positions within reference band
    #
    # observations occupy 70% of the available
    # reference band (±35% around the center)
    obs_offset = if (n() == 1) {
      
      0
      
    } else {
      
      (
        (obs_order - 1) / (n_entry - 1) - 0.5
      ) * 0.7 * first(reference_height)
      
    },
    
    # final y position
    obs_y = reference_center + obs_offset
    
  ) %>%
  
  ungroup() %>%
  
  mutate(
    realm = factor(
      realm,
      levels = c("terrestrial", "marine"),
      labels = c("Terrestrial", "Marine")
    ),
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


# ============================================== #
# Reference labels
#
# Because each reference now has its own
# equal-height band, simply use reference_center
# as the label position.


publication_labels <- rt_temp %>%
  
  group_by(
    realm,
    publication_id
  ) %>%
  
  summarise(
    label_y = first(reference_center),
    n_entry = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    publication_label = as.character(publication_id),
    
    publication_label = case_when(
      
      # References containing "et al."
      stringr::str_detect(publication_label, " et al\\. ") ~
        stringr::str_replace(
          publication_label,
          "^(.*?) et al\\. (.*)$",
          "'\\1 '*italic('et al.')*' \\2'"
        ),
      
      # References without "et al."
      TRUE ~ paste0(
        "'",
        publication_label,
        "'"
      )
    ),
    
    # Add italic (n = x)
    publication_label = paste0(
      publication_label,
      "*' ('*italic(n == ",
      n_entry,
      ")*')'"
    )
  )


# ============================================== #
# Publication colors by realm
# remove lightest 50% of each palette

publication_colors_df <- rt_temp %>%
  distinct(
    realm,
    publication_id
  ) %>%
  group_by(realm) %>%
  arrange(publication_id, .by_group = TRUE) %>%
  mutate(
    color = case_when(
      
      realm == "Marine" ~ {
        
        cols <- grDevices::hcl.colors(
          ceiling(n() / 0.65),
          palette = "Blues 3"
        )
        
        # keep darkest 50%
        cols[seq_len(n())]
      },
      
      realm == "Terrestrial" ~ {
        
        cols <- grDevices::hcl.colors(
          ceiling(n() / 0.65),
          palette = "YlOrRd"
        )
        
        # keep darkest 50%
        cols[seq_len(n())]
      },
      
      TRUE ~
        grDevices::hcl.colors(
          n(),
          palette = "Dark 3"
        )
    )
  ) %>%
  ungroup()


publication_colors <- publication_colors_df$color
names(publication_colors) <- publication_colors_df$publication_id



# ============================================== #
# Visualize publication colors

g <- ggplot(
  publication_colors_df,
  aes(
    x = publication_id,
    y = realm,
    fill = color
  )
) +
  
  geom_tile(
    color = "white",
    linewidth = 0.5
  ) +
  
  scale_fill_identity() +
  
  facet_wrap(
    ~ realm,
    scales = "free_x"
  ) +
  
  labs(
    x = "Publication",
    y = NULL
  ) +
  
  theme_bw(base_size = 18) +
  
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5
    ),
    strip.background = element_rect(fill = "grey90")
  )


# ============================================== #
# Plot

ggplot(
  rt_temp,
  aes(
    y = obs_y,
    color = publication_id
  )
) +
  
  # ---------------------------------------------- #
  # Study period
  
  geom_segment(
    aes(
      x = startperiod_startyear,
      xend = endperiod_endyear,
      yend = obs_y,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at first survey year
  
  geom_text(
    aes(
      x = startperiod_startyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at final survey year
  
  geom_text(
    aes(
      x = endperiod_endyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Reference labels
  
  geom_text(
    data = publication_labels,
    aes(
      x = -Inf,
      y = label_y,
      label = publication_label
    ),
    inherit.aes = FALSE,
    parse = TRUE,
    hjust = 1.05,
    size = 5
  ) +
  
  
  # ============================================== #
  # Scales
  
  scale_linetype_manual(
    values = c(
      "2 censuses" = "longdash",
      ">2 censuses" = "solid"
    )
  ) +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  scale_x_continuous(
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  
  scale_y_reverse(
    limits = c(1.02, -0.02),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # Equal-height realm facets
  
  facet_grid2(
    rows = vars(realm),
    scales = "fixed",
    space = "fixed"
  ) +
  
  
  coord_cartesian(
    clip = "off"
  ) +
  
  
  labs(
    x = "Year of record",
    y = NULL
  ) +
  
  
  guides(
    color = "none",
    linetype = "none"
  ) +
  
  
  theme_bw(base_size = 18) +
  
  theme(
    legend.position = "none",
    
    strip.placement = "outside",
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    panel.spacing.y = unit(0.15, "lines"),
    
    plot.margin = margin(
      t = 5.5,
      r = 5.5,
      b = 5.5,
      l = 220
    )
  )


```


## Magnitude of shift

### rt_temp

```{r}

rt_temp <- rt_cleaned %>%
  filter(
    gradient != "na",
    !is.na(magnitude_of_shift)
  ) %>%
  arrange(publication_year, publication_id) %>%
  mutate(
    gradient = stringr::str_to_sentence(gradient),
    kingdom = stringr::str_to_sentence(kingdom),
    kingdom = factor(
      kingdom,
      levels = c(
        "Chromista",
        "Plantae",
        "Animalia"
      )
    ),
    
    realm = stringr::str_to_sentence(realm),
    realm = factor(
      realm,
      levels = c(
        "Terrestrial",
        "Marine"
      )
    ),
    
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  )

```


### 1: facet plot

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 10
#| fig-height: 5


# ============================================== #
# per ecosystem x magnitude x gradient

ggplot(rt_temp, aes(x = realm, y = magnitude_of_shift)) +
  
  # horizontal reference line
  geom_hline(
    yintercept = 0,
    color = "black",
    linewidth = 0.6
  ) +
  # geom_boxplot(
  #   fill = "grey70",
  #   color = "black",
  #   outlier.shape = NA
  # ) +
  geom_jitter(
    aes(color = publication_id),
    width = 0.2,
    alpha = 0.8
  ) +
  facet_grid2(
    . ~ gradient,
    scales = "free",
    independent = "y"
  ) +
  labs(
    x = "Realm",
    y = "Magnitude of shift",
    color = "Reference no",
    title = "Shift magnitude across realms and gradients"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### 2: patchwork 1

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 10
#| fig-height: 5


# ============================================== #
# common plotting function

kingdom_colors <- c(
  "Animalia" = "#0072B2",
  "Plantae" = "#009E73",
  "Chromista" = "#D55E00"
)

kingdom_shapes <- c(
  "Animalia" = 16,
  "Plantae" = 17,
  "Chromista" = 15
)



make_shift_plot <- function(data, gradient_value, y_title) {
  
  ggplot(
    data %>% filter(gradient == gradient_value),
    aes(
      x = realm,
      y = magnitude_of_shift,
      shape = kingdom,
      color = kingdom
    )
  ) +
    
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.6
    ) +
    
    geom_jitter(
      width = 0.2,
      alpha = 0.8
    ) +
    
    scale_shape_manual(
      values = kingdom_shapes
    ) +
    
    scale_color_manual(
      values = kingdom_colors
    ) +
    
    labs(
      x = NULL,
      y = y_title
    ) +
    
    theme_bw(base_size = 16) +
    
    theme(
      legend.position = "none"
    )
}


# ============================================== #
# Create manual kingdom legend

legend_data <- tibble(
  kingdom = factor(
    c(
      "Animalia",
      "Plantae",
      "Chromista"
    ),
    levels = c(
      "Animalia",
      "Plantae",
      "Chromista"
    )
  ),
  x = 1,
  y = 1
)


legend_plot <- ggplot(
  legend_data,
  aes(
    x = x,
    y = y,
    shape = kingdom,
    color = kingdom
  )
) +
  
  geom_point(
    size = 3
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  labs(
    shape = "Kingdom",
    color = "Kingdom"
  ) +
  
  theme_void(base_size = 16) +
  
  theme(
    legend.position = "top"
  )


kingdom_legend <- cowplot::get_legend(
  legend_plot
)




# ============================================== #
# per ecosystem x magnitude x gradient

g1 <- make_shift_plot(
  rt_temp,
  gradient_value = "Elevation",
  y_title = "Elevational shift (m/year)"
)

g2 <- make_shift_plot(
  rt_temp,
  gradient_value = "Latitude",
  y_title = "Latitudinal shift (km/year)"
)

g3 <- make_shift_plot(
  rt_temp,
  gradient_value = "Longitude",
  y_title = "Longitudinal shift (km/year)"
)

g4 <- make_shift_plot(
  rt_temp,
  gradient_value = "Horizontal",
  y_title = "Horizontal shift (km/year)"
)

g_main <- g1 + g2 + g3 + g4 +
  plot_layout(
    nrow = 1
  )



# ============================================== #
# combine

g <- patchwork::wrap_elements(
  kingdom_legend
) /
  g_main +
  plot_layout(
    heights = c(0.08, 1)
  )

g







```



### 3: patchwork 2

```{r}
#| eval: FALSE
#| echo: FALSE


# ============================================== #
# common plotting function

kingdom_colors <- c(
  "Animalia" = "#0072B2",
  "Plantae" = "#009E73",
  "Chromista" = "#D55E00"
)

kingdom_shapes <- c(
  "Animalia" = 16,
  "Plantae" = 17,
  "Chromista" = 15
)


make_shift_plot <- function(data, gradient_value, y_title) {
  
  ggplot(
    data %>% filter(gradient == gradient_value),
    aes(
      x = 1,
      y = magnitude_of_shift,
      shape = kingdom,
      color = kingdom
    )
  ) +
    
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.6
    ) +
    
    geom_jitter(
      width = 0.05,
      height = 0,
      alpha = 0.8,
      size = 3
    ) +
    
    scale_shape_manual(
      values = kingdom_shapes
    ) +
    
    scale_color_manual(
      values = kingdom_colors
    ) +
    
    facet_wrap(
      ~ realm,
      nrow = 2
    ) +
    
    scale_x_continuous(
      breaks = NULL
    ) +
    
    labs(
      x = NULL,
      y = y_title
    ) +
    
    theme_bw(base_size = 20) +
    
    theme(
      legend.position = "none",
      axis.ticks.x = element_blank(),
      strip.background = element_rect(fill = "grey90")
    )
}

# ============================================== #
# Create manual kingdom legend

legend_data <- tibble(
  kingdom = factor(
    c(
      "Animalia",
      "Plantae",
      "Chromista"
    ),
    levels = c(
      "Animalia",
      "Plantae",
      "Chromista"
    )
  ),
  x = 1,
  y = 1
)


legend_plot <- ggplot(
  legend_data,
  aes(
    x = x,
    y = y,
    shape = kingdom,
    color = kingdom
  )
) +
  
  geom_point(
    size = 3
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  labs(
    shape = "Kingdom",
    color = "Kingdom"
  ) +
  
  theme_void(base_size = 20) +
  
  theme(
    legend.position = "top"
  )


kingdom_legend <- cowplot::get_legend(
  legend_plot
)




# ============================================== #
# per ecosystem x magnitude x gradient

g1 <- make_shift_plot(
  rt_temp,
  gradient_value = "Elevation",
  y_title = "Elevational shift (m/year)"
)

g2 <- make_shift_plot(
  rt_temp,
  gradient_value = "Latitude",
  y_title = "Latitudinal shift (km/year)"
)

g3 <- make_shift_plot(
  rt_temp,
  gradient_value = "Longitude",
  y_title = "Longitudinal shift (km/year)"
)

g4 <- make_shift_plot(
  rt_temp,
  gradient_value = "Horizontal",
  y_title = "Horizontal shift (km/year)"
)

g_main <- g1 + g2 + g3 + g4 +
  plot_layout(
    nrow = 1
  )



# ============================================== #
# combine

g <- patchwork::wrap_elements(
  kingdom_legend
) /
  g_main +
  plot_layout(
    heights = c(0.08, 1)
  )

g







```


### 4: patchwork 3

```{r}
#| eval: FALSE
#| echo: FALSE


# ============================================== #
# common plotting function

kingdom_colors <- c(
  "Animalia" = "#0072B2",
  "Plantae" = "#009E73",
  "Chromista" = "#D55E00"
)

kingdom_shapes <- c(
  "Animalia" = 16,
  "Plantae" = 17,
  "Chromista" = 15
)


make_shift_plot <- function(data, gradient_value, y_title) {
  
  d <- data %>%
    filter(
      gradient == gradient_value
    ) %>%
    group_by(
      realm,
      gradient,
      kingdom
    ) %>%
    mutate(
      n_group = n()
    ) %>%
    ungroup()
  
  
  # groups with enough observations for violin
  d_violin <- d %>%
    filter(
      n_group > 5
    )
  
  
  ggplot(
    d,
    aes(
      x = kingdom,
      y = magnitude_of_shift
    )
  ) +
    
    # ---------------------------------------------- #
    # zero line
    
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.6
    ) +
    
    
    # ---------------------------------------------- #
    # violin only for groups with > 5 observations
    
    geom_violin(
      data = d_violin,
      aes(
        fill = kingdom,
        color = kingdom
      ),
      width = 0.8,
      alpha = 0.25,
      linewidth = 0.8,
      trim = FALSE
    ) +
    
    
    # ---------------------------------------------- #
    # show points for ALL observations
    
    geom_jitter(
      aes(
        shape = kingdom,
        color = kingdom
      ),
      width = 0.08,
      height = 0,
      size = 3,
      alpha = 0.8
    ) +
    
    
    # ============================================== #
    # scales
    
    scale_shape_manual(
      values = kingdom_shapes
    ) +
    
    scale_color_manual(
      values = kingdom_colors
    ) +
    
    scale_fill_manual(
      values = kingdom_colors
    ) +
    
    
    # ============================================== #
    # facets
    
    facet_wrap(
      ~ realm,
      nrow = 2
    ) +
    
    
    labs(
      x = NULL,
      y = y_title
    ) +
    
    
    theme_bw(base_size = 20) +
    
    theme(
      legend.position = "none",
      strip.background = element_rect(
        fill = "grey90"
      )
    )
}

# ============================================== #
# Create manual kingdom legend

legend_data <- tibble(
  kingdom = factor(
    c(
      "Animalia",
      "Plantae",
      "Chromista"
    ),
    levels = c(
      "Animalia",
      "Plantae",
      "Chromista"
    )
  ),
  x = 1,
  y = 1
)


legend_plot <- ggplot(
  legend_data,
  aes(
    x = x,
    y = y,
    shape = kingdom,
    color = kingdom
  )
) +
  
  geom_point(
    size = 3
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  labs(
    shape = "Kingdom",
    color = "Kingdom"
  ) +
  
  theme_void(base_size = 20) +
  
  theme(
    legend.position = "top"
  )


kingdom_legend <- cowplot::get_legend(
  legend_plot
)




# ============================================== #
# per ecosystem x magnitude x gradient

g1 <- make_shift_plot(
  rt_temp,
  gradient_value = "Elevation",
  y_title = "Elevational shift (m/year)"
)

g2 <- make_shift_plot(
  rt_temp,
  gradient_value = "Latitude",
  y_title = "Latitudinal shift (km/year)"
)

g3 <- make_shift_plot(
  rt_temp,
  gradient_value = "Longitude",
  y_title = "Longitudinal shift (km/year)"
)

g4 <- make_shift_plot(
  rt_temp,
  gradient_value = "Horizontal",
  y_title = "Horizontal shift (km/year)"
)

g_main <- g1 + g2 + g3 + g4 +
  plot_layout(
    nrow = 1
  )



# ============================================== #
# combine

g <- patchwork::wrap_elements(
  kingdom_legend
) /
  g_main +
  plot_layout(
    heights = c(0.08, 1)
  )

g








```


### 5: patchwork 4

```{r}
#| eval: FALSE
#| echo: FALSE
#| fig-width: 16
#| fig-height: 10


# ============================================== #
# common plotting function

kingdom_colors <- c(
  "Animalia" = "#0072B2",
  "Plantae" = "#009E73",
  "Chromista" = "#D55E00"
)

kingdom_shapes <- c(
  "Animalia" = 16,
  "Plantae" = 17,
  "Chromista" = 15
)


make_shift_plot <- function(data, gradient_value, y_title) {
  
  d <- data %>%
    filter(
      gradient == gradient_value
    ) %>%
    group_by(
      realm,
      gradient,
      kingdom
    ) %>%
    mutate(
      n_group = n()
    ) %>%
    ungroup()
  
  
  # groups with enough observations for violin
  d_violin <- d %>%
    filter(
      n_group > 5
    )
  
  
  ggplot(
    d,
    aes(
      x = kingdom,
      y = magnitude_of_shift
    )
  ) +
    
    # ---------------------------------------------- #
    # zero line
    
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.6
    ) +
    
    
    # ---------------------------------------------- #
    # violin only for groups with > 5 observations
    
    geom_violin(
      data = d_violin,
      aes(
        fill = kingdom,
        color = kingdom
      ),
      width = 0.8,
      alpha = 0.25,
      linewidth = 0.8,
      trim = FALSE
    ) +
    
    
    # ---------------------------------------------- #
    # show points for ALL observations
    
    geom_jitter(
      aes(
        shape = kingdom,
        color = kingdom
      ),
      width = 0.08,
      height = 0,
      size = 3,
      alpha = 0.8
    ) +
    
    
    # ============================================== #
    # scales
    
    scale_shape_manual(
      values = kingdom_shapes
    ) +
    
    scale_color_manual(
      values = kingdom_colors
    ) +
    
    scale_fill_manual(
      values = kingdom_colors
    ) +
    
    
    # ============================================== #
    # facets
    
    facet_wrap(
      ~ realm,
      nrow = 2
    ) +
    
    
    labs(
      x = NULL,
      y = y_title
    ) +
    
    
    theme_bw(base_size = 20) +
    
    theme(
      legend.position = "none",
      strip.background = element_rect(
        fill = "grey90"
      ),
      axis.text.x = element_text(
        angle = 20,
        hjust = 1
      )
    )
}

# ============================================== #
# Create manual kingdom legend

legend_data <- tibble(
  kingdom = factor(
    c(
      "Animalia",
      "Plantae",
      "Chromista"
    ),
    levels = c(
      "Animalia",
      "Plantae",
      "Chromista"
    )
  ),
  x = 1,
  y = 1
)


legend_plot <- ggplot(
  legend_data,
  aes(
    x = x,
    y = y,
    shape = kingdom,
    color = kingdom
  )
) +
  
  geom_point(
    size = 3
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  labs(
    shape = "Kingdom",
    color = "Kingdom"
  ) +
  
  theme_void(base_size = 20) +
  
  theme(
    legend.position = "top"
  )


kingdom_legend <- cowplot::get_legend(
  legend_plot
)





# ============================================== #
# per ecosystem x magnitude x gradient

g1 <- make_shift_plot(
  rt_temp,
  gradient_value = "Elevation",
  y_title = "Elevational shift (m/year)"
)

g2 <- make_shift_plot(
  rt_temp,
  gradient_value = "Latitude",
  y_title = "Latitudinal shift (km/year)"
)

g3 <- make_shift_plot(
  rt_temp,
  gradient_value = "Longitude",
  y_title = "Longitudinal shift (km/year)"
)

g4 <- make_shift_plot(
  rt_temp,
  gradient_value = "Horizontal",
  y_title = "Horizontal shift (km/year)"
)





# ============================================== #
# Horizontal shift circular plot 2
# maximum = 20 km/year


# marine
rt_horizontal_20ma <- rt_temp %>%
  filter(
    gradient == "Horizontal",
    realm == "Marine",
    !is.na(azimuth_of_shift),
    !is.na(magnitude_of_shift),
    magnitude_of_shift <= 20
  )


g_horizontal_circular_20ma <- ggplot(
  rt_horizontal_20ma,
  aes(
    x = azimuth_of_shift,
    y = magnitude_of_shift,
    color = kingdom,
    shape = kingdom
  )
) +
  
  # ---------------------------------------------- #
  # movement vector
  
  geom_segment(
    aes(
      xend = azimuth_of_shift,
      y = 0,
      yend = magnitude_of_shift
    ),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  
  # ---------------------------------------------- #
  # endpoint
  
  geom_point(
    size = 3,
    alpha = 0.8
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 360),
    breaks = c(
      0, 45, 90, 135,
      180, 225, 270, 315
    ),
    labels = c(
      "N", "NE", "E", "SE",
      "S", "SW", "W", "NW"
    ),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    expand = c(0, 0)
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  
  # ============================================== #
  # circular coordinates
  
  coord_polar(
    theta = "x",
    start = 0,
    direction = 1
  ) +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  theme_bw(base_size = 20) +
  
  theme(
    legend.position = "none",
    
    plot.background = element_rect(
      color = "white",
      fill = "white",
      linewidth = 0.8
    ),
    
    plot.margin = margin(
      5, 5, 5, 5
    )
  )

g_horizontal_circular_20ma





# terrestrial
rt_horizontal_20te <- rt_temp %>%
  filter(
    gradient == "Horizontal",
    realm == "Terrestrial",
    !is.na(azimuth_of_shift),
    !is.na(magnitude_of_shift),
    magnitude_of_shift <= 20
  )


g_horizontal_circular_20te <- ggplot(
  rt_horizontal_20te,
  aes(
    x = azimuth_of_shift,
    y = magnitude_of_shift,
    color = kingdom,
    shape = kingdom
  )
) +
  
  # ---------------------------------------------- #
  # movement vector
  
  geom_segment(
    aes(
      xend = azimuth_of_shift,
      y = 0,
      yend = magnitude_of_shift
    ),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  
  # ---------------------------------------------- #
  # endpoint
  
  geom_point(
    size = 3,
    alpha = 0.8
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 360),
    breaks = c(
      0, 45, 90, 135,
      180, 225, 270, 315
    ),
    labels = c(
      "N", "NE", "E", "SE",
      "S", "SW", "W", "NW"
    ),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    expand = c(0, 0)
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  
  # ============================================== #
  # circular coordinates
  
  coord_polar(
    theta = "x",
    start = 0,
    direction = 1
  ) +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  theme_bw(base_size = 20) +
  
  theme(
    legend.position = "none",
    
    plot.background = element_rect(
      color = "white",
      fill = "white",
      linewidth = 0.8
    ),
    
    plot.margin = margin(
      5, 5, 5, 5
    )
  )

g_horizontal_circular_20te











# ============================================== #
# add circular plot as inset

g4_add <- g4 +
  patchwork::inset_element(
    g_horizontal_circular_20ma,
    left = 0.35,
    bottom = 0.1,
    right = 0.95,
    top = 0.5,
    align_to = "panel"
  ) +
  patchwork::inset_element(
    g_horizontal_circular_20te,
    left = 0.35,
    bottom = 0.6,
    right = 0.95,
    top = 1.1,
    align_to = "panel"
  )


g4_add



g_main <- g1 + g2 + g3 + g4_add +
  plot_layout(
    nrow = 1
  )





# ============================================== #
# combine

g <- patchwork::wrap_elements(
  kingdom_legend
) /
  g_main +
  plot_layout(
    heights = c(0.08, 1)
  )

g







```

### 6: patchwork 5

#### ele, lat, lon

```{r}
#| fig-width: 10
#| fig-height: 6


# ============================================== #
# common plotting function

kingdom_colors <- c(
  "Animalia" = "#0072B2",
  "Plantae" = "#009E73",
  "Chromista" = "#D55E00"
)

kingdom_shapes <- c(
  "Animalia" = 16,
  "Plantae" = 17,
  "Chromista" = 15
)


make_shift_plot <- function(data, gradient_value, y_title, n_row) {
  
  d <- data %>%
  filter(
    gradient == gradient_value
  ) %>%
  mutate(
    kingdom = factor(
      kingdom,
      levels = c(
        "Chromista",
        "Plantae",
        "Animalia"
      )
    )
  ) %>%
  group_by(
    realm,
    gradient,
    kingdom
  ) %>%
  mutate(
    n_group = n()
  ) %>%
  ungroup()
  
  
  # groups with enough observations for violin
  d_violin <- d %>%
    filter(
      n_group > 5
    )
  
  
  ggplot(
    d,
    aes(
      x = kingdom,
      y = magnitude_of_shift
    )
  ) +
    
    # ---------------------------------------------- #
    # zero line
    
    geom_hline(
      yintercept = 0,
      color = "black",
      linewidth = 0.6
    ) +
    
    
    # ---------------------------------------------- #
    # violin only for groups with > 5 observations
    
    geom_violin(
      data = d_violin,
      aes(
        fill = kingdom,
        color = kingdom
      ),
      width = 1,
      alpha = 0.25,
      linewidth = 0.8,
      trim = FALSE
    ) +
    
    
    # ---------------------------------------------- #
    # show points for ALL observations
    
    geom_jitter(
      aes(
        shape = kingdom,
        color = kingdom
      ),
      width = 0.2,
      height = 0,
      size = 3,
      alpha = 0.8
    ) +
    
    
    # ============================================== #
    # scales
    
    scale_x_discrete(
      limits = function(x) {
        intersect(
          c(
            "Chromista",
            "Plantae",
            "Animalia"
          ),
          x
        )
      }
    ) +
    
    
    scale_shape_manual(
      values = kingdom_shapes
    ) +
    
    scale_color_manual(
      values = kingdom_colors
    ) +
    
    scale_fill_manual(
      values = kingdom_colors
    ) +
    
    
    # ============================================== #
    # facets
    
    facet_wrap(
      ~ realm,
      nrow = n_row
    ) +
    
    
    labs(
      x = NULL,
      y = y_title
    ) +
    
    
    theme_bw(base_size = 20) +
    
    theme(
      legend.position = "none",
      strip.background = element_rect(
        fill = "grey90"
      ),
      axis.text.x = element_text(
        angle = 20,
        hjust = 1
      )
    )
}

# ============================================== #
# Create manual kingdom legend

legend_data <- tibble(
  kingdom = factor(
    c(
      "Chromista",
      "Plantae",
      "Animalia"
    ),
    levels = c(
      "Chromista",
      "Plantae",
      "Animalia"
    )
  ),
  x = 1,
  y = 1
)


legend_plot <- ggplot(
  legend_data,
  aes(
    x = x,
    y = y,
    shape = kingdom,
    color = kingdom
  )
) +
  
  geom_point(
    size = 3
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  labs(
    shape = "Kingdom",
    color = "Kingdom"
  ) +
  
  theme_void(base_size = 20) +
  
  theme(
    legend.position = "top"
  )


kingdom_legend <- cowplot::get_legend(
  legend_plot
)





# ============================================== #
# per ecosystem x magnitude x gradient

g1 <- make_shift_plot(
  rt_temp,
  gradient_value = "Elevation",
  y_title = "Elevational shift (m/year)",
  n_row = 2
)

g2 <- make_shift_plot(
  rt_temp,
  gradient_value = "Latitude",
  y_title = "Latitudinal shift (km/year)",
  n_row = 2
)

g3 <- make_shift_plot(
  rt_temp,
  gradient_value = "Longitude",
  y_title = "Longitudinal shift (km/year)",
  n_row = 2
)


g_main <- g1 + g2 + g3 + 
  plot_layout(
    nrow = 1
  )

g_main
```

#### horizontal

```{r}
#| fig-width: 10
#| fig-height: 6


# ============================================== #
# horizontal shift

g4 <- make_shift_plot(
  rt_temp,
  gradient_value = "Horizontal",
  y_title = "Horizontal shift (km/year)",
  n_row = 1
)




# ============================================== #
# Horizontal shift circular plot
# maximum = 20 km/year


# marine
rt_horizontal_20ma <- rt_temp %>%
  filter(
    gradient == "Horizontal",
    realm == "Marine",
    !is.na(azimuth_of_shift),
    !is.na(magnitude_of_shift),
    magnitude_of_shift <= 20
  )


g_horizontal_circular_20ma <- ggplot(
  rt_horizontal_20ma,
  aes(
    x = azimuth_of_shift,
    y = magnitude_of_shift,
    color = kingdom,
    shape = kingdom
  )
) +
  
  # ---------------------------------------------- #
  # movement vector
  
  geom_segment(
    aes(
      xend = azimuth_of_shift,
      y = 0,
      yend = magnitude_of_shift
    ),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  
  # ---------------------------------------------- #
  # endpoint
  
  geom_point(
    size = 3,
    alpha = 0.8
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 360),
    breaks = c(
      0, 45, 90, 135,
      180, 225, 270, 315
    ),
    labels = c(
      "N", "NE", "E", "SE",
      "S", "SW", "W", "NW"
    ),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    expand = c(0, 0)
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  
  # ============================================== #
  # circular coordinates
  
  coord_polar(
    theta = "x",
    start = 0,
    direction = 1
  ) +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  theme_bw(base_size = 20) +
  
  theme(
    legend.position = "none",
    
    plot.background = element_rect(
      color = "white",
      fill = "white",
      linewidth = 0.8
    ),
    
    plot.margin = margin(
      5, 5, 5, 5
    )
  )





# terrestrial
rt_horizontal_20te <- rt_temp %>%
  filter(
    gradient == "Horizontal",
    realm == "Terrestrial",
    !is.na(azimuth_of_shift),
    !is.na(magnitude_of_shift),
    magnitude_of_shift <= 20
  )


g_horizontal_circular_20te <- ggplot(
  rt_horizontal_20te,
  aes(
    x = azimuth_of_shift,
    y = magnitude_of_shift,
    color = kingdom,
    shape = kingdom
  )
) +
  
  # ---------------------------------------------- #
  # movement vector
  
  geom_segment(
    aes(
      xend = azimuth_of_shift,
      y = 0,
      yend = magnitude_of_shift
    ),
    linewidth = 0.5,
    alpha = 0.5
  ) +
  
  # ---------------------------------------------- #
  # endpoint
  
  geom_point(
    size = 3,
    alpha = 0.8
  ) +
  
  
  # ============================================== #
  # scales
  
  scale_x_continuous(
    limits = c(0, 360),
    breaks = c(
      0, 45, 90, 135,
      180, 225, 270, 315
    ),
    labels = c(
      "N", "NE", "E", "SE",
      "S", "SW", "W", "NW"
    ),
    expand = c(0, 0)
  ) +
  
  scale_y_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 5),
    expand = c(0, 0)
  ) +
  
  scale_shape_manual(
    values = kingdom_shapes
  ) +
  
  scale_color_manual(
    values = kingdom_colors
  ) +
  
  
  # ============================================== #
  # circular coordinates
  
  coord_polar(
    theta = "x",
    start = 0,
    direction = 1
  ) +
  
  labs(
    x = NULL,
    y = NULL
  ) +
  
  theme_bw(base_size = 20) +
  
  theme(
    legend.position = "none",
    
    plot.background = element_rect(
      color = "white",
      fill = "white",
      linewidth = 0.8
    ),
    
    plot.margin = margin(
      5, 5, 5, 5
    )
  )




# g_horizontal_circular_20ma
# g_horizontal_circular_20te






# ============================================== #
# add circular plot as inset

g4_add <- g4 +
  patchwork::inset_element(
    g_horizontal_circular_20te,
    left = 0.1,
    right = 0.4,
    bottom = 0.48,
    top = 0.89,
    align_to = "full"
  ) +
  patchwork::inset_element(
    g_horizontal_circular_20ma,
    left = 0.55,
    right = 0.85,
    bottom = 0.48,
    top = 0.89,
    align_to = "full"
  )


g4_add



```


## shift_tested

```{r}
#| fig-width: 10
#| fig-height: 5

# ============================================== #
# count rows per ecosystem x shift_tested x gradient

rt_temp <- rt_cleaned %>%
  filter(gradient != "na") %>%
  count(
    gradient,
    realm,
    shift_tested,
    realm
  )


ggplot(
  rt_temp,
  aes(
    x = realm,
    y = n,
    fill = shift_tested
  )
) +
  geom_col(position = "stack") +
  facet_grid2(
    . ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  labs(
    x = "Realm",
    y = "Number of observations",
    fill = "Shift tested"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )
```


```{r}
#| fig-width: 10
#| fig-height: 5


# ============================================== #
# bar plot: count rows per ecosystem x shift_significant x gradient

rt_temp <- rt_cleaned %>%
  filter(
    gradient != "na",
    shift_tested == "yes"
  ) %>%
  count(
    gradient,
    realm,
    shift_significant,
    realm
  )


ggplot(
  rt_temp,
  aes(
    x = realm,
    y = n,
    fill = shift_significant
  )
) +
  geom_col(position = "stack") +
  facet_grid2(
    . ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  labs(
    x = "Realm",
    y = "Number of observations",
    fill = "Shift significant"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )






```





# Session info

```{r}
#| label: session-info

sessioninfo::session_info()
```

```{r}
#| label: run-info
#| echo: false

time_end <- Sys.time()
elapsed_sec <- as.numeric(difftime(time_end, time_start, units = "secs"))
elapsed_min <- as.numeric(difftime(time_end, time_start, units = "mins"))

readr::write_lines(
  c(
    paste("rendered_at:", Sys.time()),
    paste("input_file:", in1),
    paste("input_file:", in2),
    paste("output_dir:", dir_out),
    paste("run_time_seconds:", round(elapsed_sec, 1)),
    paste("run_time_minutes:", round(elapsed_min, 2))),
  file.path(dir_out, "run_info.txt")
)
```
