---
title: "reviewtable_organize"
author: "tksym"
date: today
params:
  in1: null
  in2: null
  dir_out: null
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: right
    fig-format: png
    embed-resources: true
    self-contained-math: true
    code-fold: true
    code-overflow: wrap
  pdf:
    toc: false
editor: 
  mode: source
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  fig.width = 16,
  fig.height = 9,
  dpi = 120,
  out.width = "100%",
  fig.align = "center",
  cache = FALSE
)

time_start <- Sys.time()
```

```{r}
#| label: packages

pacman::p_load(
  # data import
  here,
  fs,
  readxl,

  # data manipulation
  tidyverse,
  
  # plot
  ggh4x,

  # misc
  tictoc,
  sessioninfo,
  report
)

```

```{r}
#| label: othersettings
#| include: false

# assign from outside
in1 <- params$in1
in2 <- params$in2
dir_out <- params$dir_out


```

# WORK FLOW

- Import data

- Check original data

- Data cleaning

- Visualize number of range shift observations per ecosystem

# DATA

Importing previous review table: `{r} fs::path_rel(in1)`

Importing updated review table: `{r} fs::path_rel(in1)`

Output directory: `{r} fs::path_rel(dir_out)`



# Import

```{r}

# original data
rt_prev <- read_csv(in1)


# data after small checks in excel
rt <- read_xlsx(in2, sheet = "dataentry_ok")

# other sheets
lon_latrange <- read_xlsx(in2, sheet = "longitude_latrange")
ref_en <- read_xlsx(in2, sheet = "En-list_ok")
ref_ja <- read_xlsx(in2, sheet = "Ja-list_ok")




```



# Check original data

```{r}
#| eval: FALSE
#| echo: FALSE


# glimpse(rt_prev)

# ============================================== #
# vars for simple cleaning

cols_upper_prev <- c("recorder_name",
                "reference_no",
                "study_region",
                "area_name",
                "taxonomic_unit",
                "taxonomic_unit_revised")




# ============================================== #
# simple cleaning

rt_prev2 <- rt_prev %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  
  # convert text to lower (except specified columns)
  mutate(across(c(where(is.character),
                  -all_of(cols_upper_prev)),
                str_to_lower)) %>% 
  
  # replace NA with "na"
  mutate(across(everything(),
                ~ tidyr::replace_na(.x, "na")))




# glimpse(rt_prev2)
```


```{r}
#| eval: FALSE
#| echo: FALSE

# ============================================== #
# check: show unique values for each column

nrow(rt_prev2)

for (nm in names(rt_prev2)) {
  vals <- unique(rt_prev2[[nm]])

  cat("\n====================\n")
  cat("Column:", nm, "\n")
  cat("n_unique:", length(vals), "\n")
  cat("====================\n")

  print(vals)

}


```







# Clean range shfit data 1 (light)

```{r}
# ============================================== #
# simple cleaning

# glimpse(rt)
# glimpse(rt2)


cols_upper <- c("recorder_name",
                "reference_no",
                "study_area",
                "specific_area",
                "taxonomic_unit",
                "taxonomic_unit_revised")



rt2 <- rt %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  
  # convert text to lower (except specified columns)
  mutate(across(c(where(is.character),
                  -all_of(cols_upper)),
                str_to_lower)) %>% 
  
  # replace NA with "na"
  mutate(across(everything(),
                ~ tidyr::replace_na(.x, "na")))

# glimpse(rt2)



# ============================================== #
# filter out irrelevant data at initial stage
# edit start period and end period data

rt2$include %>% unique() %>% sort()

rt3 <- rt2 %>% 
  filter(
    # this include col was made in excel for rows based on:
    # n_years: too large (comparison with glacial time)
    # gradient = na
    # magnitude_of_shift = na
    # notes = no resurvey
    # include == 2 is "pending" literature
    include %in% 1) %>%
  
  # start period
  rename(startperiod_orig = start_period) %>%
  mutate(
    # extract all 4-digit years
    startperiod_years = str_extract_all(startperiod_orig, "\\d{4}"),
    
    startperiod_startyear = map_int(
      startperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[1]))
    ),
    
    startperiod_endyear = map_int(
      startperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[length(.x)]))
    ),
    
    startperiod_midyear = (startperiod_startyear + startperiod_endyear) / 2
  ) %>%
  select(-startperiod_years) %>% 
  relocate(startperiod_startyear, startperiod_endyear, startperiod_midyear, .after = startperiod_orig) %>% 
  
  # end period
  rename(endperiod_orig = end_period) %>%
  mutate(
    # extract all 4-digit years
    endperiod_years = str_extract_all(endperiod_orig, "\\d{4}"),
    
    endperiod_startyear = map_int(
      endperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[1]))
    ),
    
    endperiod_endyear = map_int(
      endperiod_years,
      ~ ifelse(length(.x) == 0, NA_integer_, as.integer(.x[length(.x)]))
    ),
    
    endperiod_midyear = (endperiod_startyear + endperiod_endyear) / 2
  ) %>%
  select(-endperiod_years) %>% 
  relocate(endperiod_startyear, endperiod_endyear, endperiod_midyear, .after = endperiod_orig) %>% 
  
  # n_years
  mutate(
    n_years_midyear = endperiod_midyear - startperiod_midyear
  ) %>% 
  relocate(n_years_midyear, .after = n_years)

nrow(rt2)
nrow(rt3)
```




# Clean reference info
```{r}

# ============================================== #
# minor edits

# glimpse(ref_en)
# glimpse(ref_ja)

ref_en2 <- ref_en %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower())

ref_ja2 <- ref_ja %>% 
  # convert space in columns to underbar
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower())


# glimpse(ref_en2)
# glimpse(ref_ja2)


ref_en3 <- ref_en2 %>% 
  filter(reference_no %in% unique(rt3$reference_no)) %>% 
  select(reference_no, article_title, data_section, source_title, publication_year, authors, doi) %>% 
  mutate(publication_year = as.numeric(publication_year),
         language = "English")

ref_ja3 <- ref_ja2 %>% 
  filter(reference_no %in% unique(rt3$reference_no)) %>% 
  select(reference_no, article_title, data_section, source_title, publication_year, authors, doi) %>% 
  mutate(language = "Japanese")

# glimpse(ref_en3)
# glimpse(ref_ja3)



# ============================================== #
# bind
ref_bind <- bind_rows(ref_en3,
                      ref_ja3) %>% 
  relocate(language, reference_no, doi, authors, article_title, publication_year, source_title, data_section) %>% 
  mutate(author_year = str_c(authors, publication_year, sep = " "))


# ============================================== #
# check and change cell entries


# ------------ #
ref_bind$author_year %>% unique()
ref_bind$author_year %>% class()

authors_recode <- c(
  "Sakio & Masuzawa 2012" = "Sakio & Masuzawa 2012",
  "Jung, S; Pang, IC; Lee, JH; Choi, I; Cha, HK 2014" = "Jung et al. 2014",
  "Winkler, DE; Amagai, Y; Huxman, TE; Kaneko, M; Kudo, G 2016" = "Winkler et al. 2016",
  "Shimazaki, M; Sasaki, T; Hikosaka, K; Nakashizuka, T 2011" = "Shimazaki et al. 2011",
  "Tougou, D; Musolin, DL; Fujisaki, K 2009" = "Tougou et al. 2009",
  "Kakehi, S; Narimatsu, Y; Okamura, Y; Yagura, A; Ito, S 2021" = "Kakehi et al. 2021",
  "Kurihara, T; Takami, H; Kosuge, T; Chiba, S; Iseda, M; Sasaki, T 2011" = "Kurihara et al. 2011",
  "T Sakiyama 2024" = "Sakiyama 2024",
  "Naoki H. Kumagai, Jorge García Molinosa, Hiroya Yamano, Shintaro Takao, Masahiko Fujii, Yasuhiro Yamanaka 2018" = "Kumagai et al. 2018",
  "S. Watari, S. Takemura, H. Oyaizu 2024" = "Watari et al. 2024",
  "Tu, CY; Tian, YJ; Hsieh, CH 2015" = "Tu et al. 2015",
  "Yamano, H; Sugihara, K; Nomura, K 2011" = "Yamano et al. 2011",
  "Y Kawauchi, Y Yagi, T Yano, K Fujiwara 2023" = "Kawauchi et al. 2023",
  "K Higashiguchi, S Nishijima, Y Kamimura… 2025" = "Higashiguchi et al. 2025",
  "Y Kawakami, K Yamazaki, K Ohashi 2014" = "Kawakami et al. 2014",
  "山内 貴義, 工藤 雅志, 高槻 成紀 2007" = "Yamauchi et al. 2007",
  "北原 正彦 2008" = "Kitahara 2008",
  "木澤 遼; 濱 侃; 吉田 圭一郎 2022" = "Kizawa et al. 2022",
  "奥村 忠誠; 清水 庸; 大政 謙次 2009" = "Okumura et al. 2009"
)


# ------------ #
ref_bind$source_title %>% unique()

journal_recode <- c(
  "Journal of Plant Research" = "Journal of Plant Research",
  "Reviews in fish biology and fisheries" = "Reviews in Fish Biology and Fisheries",
  "Plant ecology" = "Plant Ecology",
  "Global change biology" = "Global Change Biology",
  "Entomologia experimentalis et applicata" = "Entomologia Experimentalis et Applicata",
  "Marine ecology progress series" = "Marine Ecology Progress Series",
  "Marine biology" = "Marine Biology",
  "PhD dissertation" = "PhD dissertation",
  "PNAS" = "PNAS",
  "Frontiers in Marine Science" = "Frontiers in Marine Science",
  "Fisheries oceanography" = "Fisheries Oceanography",
  "Geophysical research letters" = "Geophysical Research Letters",
  "Regional studies in marine science" = "Regional Studies in Marine Science",
  "SSRN" = "SSRN",
  "Applied entomology and zoology" = "Applied Entomology and Zoology",
  "哺乳類科学" = "哺乳類科学",
  "昆虫と自然" = "昆虫と自然",
  "E-journal GEO" = "E-journal GEO",
  "環境科学会誌" = "環境科学会誌"
)


# ------------ #

ref_bind2 <- ref_bind %>% 
  mutate(publication_id = recode(author_year, !!!authors_recode)) %>% 
  relocate(publication_id, .after = reference_no) %>% 
  select(-author_year) %>%
  mutate(
    source_title = recode(source_title, !!!journal_recode)
  ) %>% 
  arrange(language, publication_year)





```



# Clean range shfit data 2 (heavy)

```{r}
# ============================================== #
# column names

names(rt3)



# ============================================== #
# check and change cell entries


# ------------ #
rt3$geographic_scale %>% unique()

geographicscale_recode <- c()


# ------------ #
rt3$study_area %>% unique()

studyarea_recode <- c()


# ------------ #
rt3$specific_area %>% unique()

specificarea_recode <- c()


# ------------ #
rt3$kingdom %>% unique()

kingdom_recode <- c(
  "plant" = "plantae",
  "animal" = "animalia",
  "chromista" = "chromista",
  "動物" = "animalia",
  "植物" = "plantae"
  
)

kingdom_recode %>% unique()

# ------------ #
rt3$taxonomic_resolution %>% unique()

taxres_recode <- c(
  "species-level" = "species",
  "species" = "species",
  "種レベル" = "species",
  "order-level" = "order")

taxres_recode %>% unique()


# ------------ #
rt3$taxonomic_unit %>% unique()

# recode taxonomic_unit_revised
taxunit_recode <- c()


# ------------ #
rt3$taxonomic_unit_revised %>% unique() %>% sort()

# supported by ai
taxunitrev_recode <- c(

  # check later: na
  "na" = "na"
)

taxunitrev_recode %>% unique()


# ------------ #
rt3$ecosystem %>% unique()


ecosystem_recode <- c(
  
  # alpine
  "alpine" = "alpine",
  
  # forest
  "forest" = "forest",
  "森林" = "forest",
  "forest, alpine" = "forest", # Hitomi E0267 Hakkoda, include subalpine as forest
  
  # grassland
  "grassland" = "grassland",
  
  # wetland
  "wetlands" = "wetland",
  
  # marine
  "marine" = "marine",
  
  
  
  # check
  
  # mixed ecosystems as terrestrial
  "forest, grassland" = "terrestrial", # Dai E0400 Nezara viridula kamemushi
  "森林、草地" = "terrestrial", # Nonaka J0202 sika
  "限定なし" ="terrestrial", # Nonaka J0201 Birds
  
  # broad terrestrial kept as terrestrial
  "terrestrial" = "terrestrial" # Sueyoshi E2247 ladybug
)

ecosystem_recode %>% unique()

realm_recode <- c(
  
  # terrestrial
  "alpine" = "terrestrial",
  "forest" = "terrestrial",
  "grassland" = "terrestrial",
  "wetland" = "terrestrial",
  
  # tentative terrestrial
  "urban" = "terrestrial",

    
  # marine
  "marine" = "marine"
)


# ------------ #
rt3$information_source %>% unique()

infosrc_recode <- c(
  
  # resurvey by authors
  "resurvey" =
    "resurvey",
  "authors did the resurvey" =
    "resurvey",
  "著者が再調査を実施" =
    "resurvey",
  
  
  # remote sensing
  "remote sensing" =
    "remote sensing",
  "aerial photograph by forest agency" =
    "remote sensing",
  "geospatial information authority of japan" =
    "remote sensing",
  
  
  # literature
  "literature" =
    "literature",
  "published papers" =
    "literature",
  
  
  # literature and speciemens
  "literature and museum specimens" =
    "literature and specimens",
  "specimen records, literature" =
    "literature and specimens",
  

  # citizen and specimens
  "citizen science and specimens" =
    "citizen science and specimens",
  
  
  
  # check later with Jorge: fishery statistics, separate trawl or not?
  "fishery statistics" =
    "fishery statistics",
  "commercial fisheries data" =
    "fishery statistics",
  "fishery and aquaculture production statistics 　国の漁獲統計" =
    "fishery statistics",
  
  "bottom-trawl surveys" =
    "trawl survey",
  "trawl fishery data" =
    "trawl fishery",
  "fishery-independent trawl surveys by fisheries research and education agency (fra)" =
    "fishery-independent trawl survey",
  
  
  # check later: governmental survey
  "distribution survey by the ministry of the environment" =
    "governmental monitoring survey",
  
  
  # check later: database
  "データベースを利用" =
    "multiple sources",
  
  # check 
  "na" =
    "na" # Fatima E0608
)

infosrc_recode %>% unique()


# ------------ #
rt3$gradient %>% unique() %>% sort()

gradient_recode <- c(
  
  # elevation
  "elevation" = "elevation",
  "標高" = "elevation",
  
  # lat
  "latitude" = "latitude",
  "latitude (northward)" = "latitude",
  "latitude (poleward / northward)" = "latitude",
  
  
  # lon
  "longitude" = "longitude",
  
  
  # horizontal
  "horizontal" = "horizontal", # Jorge: I think this is ok
  "水平方向" = "horizontal", # Nonaka J0202
  "緯度・経度方向" = "horizontal" # Nonaka J0201 sika: checked and this is based on mesh data so this is horizontal
  )

gradient_recode %>% unique()



# ------------ #
rt3$position %>% unique()

position_recode <- c(
  
  # leading edge
  "leading edge" = "leading edge",
  "leading" = "leading edge",
  "northern limit" = "leading edge",
  "前縁" = "leading edge",
  
  # centroid
  "centroid" = "centroid",
  "centroid / multiple sites" = "centroid",
  
  "average" = "centroid", # Dai E0434 pika: study is about centroid
  "na" = "centroid", # Nonaka J0202 sika: authors do not mention centroid, but the study area covers entire honshu, shikoku, and kyushu, which largely covers the sika's distribution. so centroid was applied here.
  
  # need to check using distribution data, but it will be converted to centroid for now, because it is likely to reflect the centroid
  "unknown" = "centroid", # Sakiyama E0009
  
  # trailing edge
  "trailing edge" = "trailing edge",
  "trailing" = "trailing edge",
  
  
  # check
  "前縁、重心、後縁" = "others" # Nonaka J0201 bird: authors do not mention but this can be interpreted as centroid


)

position_recode %>% unique()



# ------------ #
rt3$measurement_type %>% unique()

meastype_recode <- c(
  
  # abundance
  "abundance" = "abundance",
  
  # occurrence
  "occurrence" = "occurrence",
  "在不在" = "occurrence",

  # presence-only: group as occurrence
  "occurrence (presence-only)" = "occurrence_presonly",
  
  
  
  # check
  
  # density: group as abundance
  "density from catch data" = "abundance_density",
  "density from survey" = "abundance_density",
  
  # cpue: group as abundance
  "cpue" = "abundance_cpue",
  "abundance (cpue)" = "abundance_cpue",
  
  # catch: group as abundance
  "catch landings" = "abundance_catchlandings"
  )

meastype_recode %>% unique()


observation_recode <- c(
  
  # abundance
  "abundance" = "abundance",
  "abundance_density" = "abundance",
  "abundance_cpue" = "abundance",
  "abundance_catchlandings" = "abundance",
  
  # occurrence
  "occurrence" = "occurrence",
  "occurrence_presonly" = "occurrence"
  
  
)




# ------------ #
rt3$startperiod_orig %>% unique()



# ------------ #
rt3$endperiod_orig %>% unique()



# ------------ #
rt3$n_years %>% unique()


# ------------ #
rt3$n_years_midyear %>% unique()



# ------------ #
rt3$number_of_censuses %>% unique()

ncensus_recode <- c(
  
  # check
  "na" = "na", # Harsh E0949, E1034, E_har1; Jorge E_jgm1, E1585, E1667, E1991; Fatima E0608
  
  "a total of 632 study sites over 16 years (26-59 sites per year)" = "632", # Sueyoshi E2241
  
  "the number of specimens recorded was 1312, but the number obtained from the literature was not reported." = "1312" # Sueyoshi E2247
)

ncensus_recode %>% unique()


# ------------ #
rt3$magnitude_of_shift %>% unique()

# magshift_recode <- c(
# )
# 
# magshift_recode %>% unique()


# ------------ #
rt3$unit_of_shift %>% unique()

# later
# convert cells with m, km, km/decade to km/year

unitshift_recode <- c(
  
  # meter
  "meter" = "m",
  "m" = "m",
  "メートル" = "m",
  
  # km
  "kilometer" = "km",
  "km" = "km",
  
  # km/year
  "km year-1" = "km/year",
  "km/year" = "km/year",
  
  # km/decade
  "km per decade" = "km/decade",
  
  # degree
  "degree e" = "degree",
  "degree n" = "degree",
  "degrees" = "degree",
  
  # degree/year
  "degree year-1" = "degree/year",
  "degree/year" = "degree/year",
  "degrees/year" = "degree/year"
)





# ------------ #
rt3$azimuth_of_shift %>% unique() %>% sort()

# azimuth_recode <- c(
#   
# )



# ------------ #
rt3$shift_tested %>% unique() %>% sort()

shifttest_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (centroid position)" = "yes",
  "estimated" = "yes", # Jorge E_jgm2, E1991: tested shift based on significance of coef

  # no
  "no" = "no")

# ------------ #
rt3$shift_significant %>% unique() %>% sort()

shiftsign_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (supported by consistent results across 6 modeling thresholds, here the magnitude is given based on equate entropy of thresholded and original distribution)" = "yes",

  # no
  "no" = "no",
  
  # na
  "na" = "na",
  "no analysis" = "na",
  "no analyses" = "na"
  
)

# ------------ #
rt3$data_cleaning %>% unique() %>% sort()

dataclean_recode <- c(
  
  # yes
  "yes" = "yes",
  "yes (the oldest records from each of the 39 regions were used)" =
    "yes",

  # no
  "no" = "no",
  
  # na
  "na" = "na"
  
  )

dataclean_recode %>% unique()


# ------------ #
rt3$modelling_conducted %>% unique() %>% sort()

# check again

model_recode <- c(
  
  # yes
  "yes" = "yes",
  
  "yes\r\n  (centroid shifts were estimated using vector autoregressive spatiotemporal model)" =
    "yes",
  
  # no
  "no" = "no"
  )

model_recode %>% unique()


# ------------ #
rt3$other_effects %>% unique() %>% sort()

othereff_recode <- c(
  
  # no
  "no" = "no",
  
  # na
  "na" = "na",
  
  # try to phrase in keywords
  
  "authors argue that the natural recovery after volcano eruption is the main process of the advancing elevation and discuss that climate change may accelerate this process." =
    "volcanic recovery",
  
  "human activities, species interactions, landuse" =
    "human activity, species interaction, land use",
  
  
  "①increase in population size ②decrease in snowfall due to global warming" =
    "population increase, snowfall decrease",
  
  "distance from moorland" =
    "distance from moorland",
  
  "species interactions" =
    "species interaction",
  
  "fishery pressure, species interactions" =
    "fishery pressure, species interaction",
  
  "sea current, species interaction, ultraviolet radiation, salinity stress" =
    "sea current, species interaction, ultraviolet radiation, salinity",
  
  "coastal development, deterioration of water quality" =
    "coastal development, water quality",
  
  "pacific decadal oscillation" =
    "pacific decadal oscillation",
  
  "adult dispersal ability, larval feeding range, distribution of food resources" =
    "dispersal ability, larval feeding, food resources",
  
  "land cover changes (increase in forest area due to abandonment of agricultural land)" =
    "land cover change",
  
  "precipitation, elevation, human influence index" =
    "precipitation, elevation, human influence",
  
  "土地利用変化、農地環境変化" =
    "land use change, agricultural change",
  
  "標高、広葉樹林面積、針葉樹林面積、草地面積、田んぼ面積、荒地面積、耕作放棄地面積、人口、道路延長距離、距離、積雪" =
    "elevation, land cover, population, road length, distance, snow"
)



# ------------ #
rt3$shift_source %>% unique() %>% sort()

shiftsrc_recode <- c(
  
  # stated
  "stated" = "stated",
  "stated in discussion" = "stated",
  "yes" = "stated", # 0434: case of northern pika

  # estimated
  "estimated" = "estimated"
  )

shiftsrc_recode %>% unique()


# ------------ #
rt3$notes %>% unique()

# notes_recode <- c()
```


```{r}
# ============================================== #
# clean dataframe by recoding

rt4 <- rt3 %>% 
  
  mutate(
  
  # ------------------------------------------------------------------ #
    # recode categorical columns
    
    # study_region = recode(study_region, !!!studyregion_recode),
    # area_name = recode(area_name, !!!areaname_recode),
    
    kingdom = recode(kingdom, !!!kingdom_recode),
    
    taxonomic_resolution = recode(
      taxonomic_resolution,
      !!!taxres_recode
    ),
    
    # taxonomic_unit = recode(
    #   taxonomic_unit,
    #   !!!taxunit_recode
    # ),
    
    taxonomic_unit_revised = recode(
      taxonomic_unit_revised,
      !!!taxunitrev_recode
    ),
    
    ecosystem = recode(
      ecosystem,
      !!!ecosystem_recode
    ),
    
    realm = recode(
      ecosystem,
      !!!realm_recode
    ),
    
    information_source = recode(
      information_source,
      !!!infosrc_recode
    ),
    
    gradient = recode(
      gradient,
      !!!gradient_recode
    ),
    
    position = recode(
      position,
      !!!position_recode
    ),
    
    measurement_type = recode(
      measurement_type,
      !!!meastype_recode
    ),
    
    observation = recode(
      measurement_type,
      !!!observation_recode
    ),
    
    # n_years = recode(
    #   n_years,
    #   !!!nyears_recode
    # ),
    
    number_of_censuses = recode(
      number_of_censuses,
      !!!ncensus_recode
    ),
    
    # magnitude_of_shift = recode(
    #   magnitude_of_shift,
    #   !!!magshift_recode
    # ),
    
    unit_of_shift = recode(
      unit_of_shift,
      !!!unitshift_recode
    ),
    
    # azimuth_of_shift = recode(
    #   azimuth_of_shift,
    #   !!!azimuth_recode
    # ),
    
    shift_tested = recode(
      shift_tested,
      !!!shifttest_recode
    ),
    
    shift_significant = recode(
      shift_significant,
      !!!shiftsign_recode
    ),
    
    data_cleaning = recode(
      data_cleaning,
      !!!dataclean_recode
    ),
    
    modelling_conducted = recode(
      modelling_conducted,
      !!!model_recode
    ),
    
    other_effects = recode(
      other_effects,
      !!!othereff_recode
    ),
    
    shift_source = recode(
      shift_source,
      !!!shiftsrc_recode
    ),
    
    # notes = recode(
    #   notes,
    #   !!!notes_recode
    # ),
    
    
    # ------------------------------------------------------------------ #
    # numeric conversion
    
    number_of_censuses = na_if(number_of_censuses, "na"),
    number_of_censuses = as.numeric(number_of_censuses),
    
    magnitude_of_shift = na_if(magnitude_of_shift, "na"),
    magnitude_of_shift = as.numeric(magnitude_of_shift)
    
  ) %>% 
  relocate(realm, .after = "ecosystem") %>% 
  relocate(observation, .after = "measurement_type") %>% 
  
  # ecosystem order
  mutate(
    
    # check entry with ecosystem_recode %>% unique()
    
    ecosystem = factor(
      ecosystem,
      levels = c(
        "alpine",
        "forest",
        "grassland",
        "wetland",
        "terrestrial",
        "marine"
      )
    ),
    realm = factor(
      realm,
      levels = c(
        "terrestrial",
        "marine"
        )
    ),
    gradient = factor(
      gradient,
      levels = c(
        "elevation",
        "latitude",
        "longitude",
        "horizontal")
    )
  ) %>% 
  
  left_join(ref_bind2 %>% select(-language),
            by = "reference_no")
```


```{r}
# ============================================== #
# finally edit unit


# ------------ #
# check combination of gradient and unit
rt4 %>%
  distinct(gradient, unit_of_shift) %>%
  arrange(gradient, unit_of_shift)




# ------------ #
# conversion from latitude degrees to km
latdeg_to_km <- function(deg) {
  earth_radius <- 6371
  deg * 2 * pi * earth_radius / 360
}

# ---- #
# example 

latdeg_to_km(0.1)
# ---- #



# ------------ #
# conversion from longitude degrees to km
londeg_to_km <- function(deg, latitude) {
  earth_radius <- 6371
  deg * 2 * pi * earth_radius / 360 * cos(latitude * pi / 180)
}

# ---- #
# example

londeg_to_km(1, latitude = 36)
londeg_to_km(1, latitude = 46)
# ---- #

# lat mean for longitude studies
lon_latrange2 <- lon_latrange %>% 
  rename_with(~ .x %>%
                str_replace_all(" ", "_") %>%
                str_to_lower()) %>% 
  # convert text to lower (except specified columns)
  mutate(gradient = str_to_lower(gradient)) %>% 
  filter(include == 1) %>% 
  rowwise() %>% 
  mutate(lat_mean = mean(lat_min, lat_max)) %>% 
  select(reference_no, gradient, lat_mean)
  






# ------------ #
rt5 <- rt4 %>% 
  
  left_join(lon_latrange2, 
            by = c("reference_no", "gradient")) %>% 
  
  # transform magnitudeshift and unitshift
  mutate(
    
    magnitude_of_shift = case_when(
      
      # apply these to magnitudeshift depending on
      # gradient
      # unitshift
      
      
      # ------------------------------------------------------
      # elevation --> m/year
      # ------------------------------------------------------
      
      # m --> m/year
      gradient == "elevation" & unit_of_shift == "m" ~
        magnitude_of_shift / n_years_midyear,
      
      
      # ------------------------------------------------------
      # latitude --> km/year
      # ------------------------------------------------------
      
      # degree --> km --> km/year
      gradient == "latitude" & unit_of_shift == "degree" ~
        latdeg_to_km(magnitude_of_shift) / n_years_midyear,
      
      # degree/year --> km/year
      gradient == "latitude" & unit_of_shift == "degree/year" ~
        latdeg_to_km(magnitude_of_shift),
      
      # km --> km/year
      gradient == "latitude" & unit_of_shift == "km" ~
        magnitude_of_shift / n_years_midyear,
      
      # already km/year
      gradient == "latitude" & unit_of_shift == "km/year" ~
        magnitude_of_shift,
      
      
      # ------------------------------------------------------
      # longitude --> km/year
      # ------------------------------------------------------
      
      # degree --> km --> km/year
      gradient == "longitude" & unit_of_shift == "degree" ~
        londeg_to_km(magnitude_of_shift, latitude = lat_mean) / n_years_midyear,
      
      # degree/year --> km/year
      gradient == "longitude" & unit_of_shift == "degree/year" ~
        londeg_to_km(magnitude_of_shift, latitude = lat_mean),
      
      
      
      # ------------------------------------------------------
      # horizontal --> km/year
      # ------------------------------------------------------
      
      # km --> km/year
      gradient == "horizontal" & unit_of_shift == "km" ~
        magnitude_of_shift / n_years_midyear,
      
      # already km/year
      gradient == "horizontal" & unit_of_shift == "km/year" ~
        magnitude_of_shift,
      
      
      # anything unexpected
      TRUE ~ NA_real_

    ),
    
    unit_of_shift = case_when(
      gradient == "elevation" ~ "m/year",
      gradient %in% c("latitude", "longitude", "horizontal") ~ "km/year",
      TRUE ~ NA_character_
      )
    )



# ------------ #
# check combination of gradient and unit
rt5 %>%
  distinct(gradient, unit_of_shift) %>%
  arrange(gradient, unit_of_shift)



rt_cleaned <- rt5



# ------------ #
# save

readr::write_csv(
  rt_cleaned,
  fs::path(dir_out, "rt_cleaned.csv")
)

```


# Check cleaned data

```{r}
# ============================================== #
# check: show unique values for each column

nrow(rt_cleaned)

for (nm in names(rt_cleaned)) {
  vals <- unique(rt_cleaned[[nm]])
  
  cat("\n====================\n")
  cat("Column:", nm, "\n")
  cat("n_unique:", length(vals), "\n")
  cat("====================\n")
  
  print(vals)
  
}


# check entries for shift_signif when shift_tested == yes
rt_cleaned %>%
  filter(shift_tested == "yes") %>%
  pull(shift_significant) %>%
  unique()



```



# Visualize

```{r}
#| fig-width: 12
#| fig-height: 6


# glimpse(rt_cleaned)

# ============================================== #
# bar plot: count number of studies

rt_temp <- rt_cleaned %>%
  filter(!is.na(magnitude_of_shift)) %>%
  distinct(realm, ecosystem, language, reference_no) %>%
  count(realm, ecosystem, language, name = "n_studies")

rt_temp 



```


```{r}
#| fig-width: 12
#| fig-height: 6

# ============================================== #
# bar plot: count rows per ecosystem x x gradient

rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(magnitude_of_shift)
  ) %>%
  
  count(
    gradient,
    ecosystem,
    realm
  )

ggplot(rt_temp, aes(x = realm, y = n, fill = ecosystem)) +
  geom_col(position = "stack") +
  facet_grid2(
    . ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  labs(
    x = "Realm",
    y = "Number of observations",
    fill = "Ecosystem",
    title = "Number of observations per gradient",
  ) +
  theme_bw(base_size = 16) +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )






```



## Gantt chart-like plot

### 1: row height varies with sample size

```{r}
#| fig-width: 10
#| fig-height: 8


# ============================================== #
# Gantt chart-like plot of study duration per observation

rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_midyear),
    !is.na(endperiod_midyear)
  ) %>%
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(startperiod_midyear, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_midyear,
    endperiod_midyear
  ) %>%
  
  # publication legend follows arranged order
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    ),
    
    # unique observation ID across whole dataset
    obs_id = row_number()
  )%>%
  mutate(
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


class(rt_temp$obs_id)


ggplot(rt_temp, aes(y = obs_id, color = publication_id)) +
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = endperiod_midyear,
      yend = obs_id,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = startperiod_midyear+1,
      yend = obs_id
    ),
    color = "black",
    linewidth = 1,
    alpha = 1
  ) +
  geom_segment(
    aes(
      x = endperiod_midyear-1,
      xend = endperiod_midyear,
      yend = obs_id
    ),
    color = "black",
    linewidth = 1,
    alpha = 1
  ) +
  # geom_errorbar(
  #   aes(
  #     y = obs_id,
  #     xmin = startperiod_midyear,
  #     xmax = endperiod_midyear,
  #     linetype = census_linetype
  #   ),
  #   width = 0.3,
  #   linewidth = 1,
  #   orientation = "y",
  #   alpha = 1
  # ) +
  geom_text(
    aes(
      x = endperiod_midyear,
      label = number_of_censuses
    ),
    hjust = -0.3,
    color = "black",
    size = 3.5
  ) +
  scale_linetype_manual(
    values = c(
      "2 censuses" = "dotted",
      ">2 censuses" = "solid"
    )
  ) +
  # geom_point(
  #   aes(x = startperiod_midyear),
  #   size = 2,
  #   alpha = 1
  # ) +
  # geom_point(
  #   aes(x = endperiod_midyear),
  #   size = 2,
  #   alpha = 1
  # ) +
  
  facet_grid2(
    rows = vars(realm),
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  
  scale_x_continuous(
    expand = expansion(mult = c(0.02, 0.1))
  ) +
  
  scale_y_reverse() +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  labs(
    x = "Year",
    y = "Observation",
    color = "Reference no",
    title = "Years between the original survey and resurvey"
  ) +
  
  guides(
    color = guide_legend(ncol = 1)
  ) +
  
  theme_bw(base_size = 16) +
  theme(
    legend.position = "right",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    plot.title = element_text(hjust = 0),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )


```


### 2: row height is fixed regardless of sample size

```{r}
#| fig-width: 10
#| fig-height: 8

# ============================================== #
# Gantt chart-like plot of study duration per observation
#
# Equal vertical height for:
#   1. each realm
#   2. each reference within each realm


rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_midyear),
    !is.na(endperiod_midyear)
  ) %>%
  
  
  # ---------------------------------------------- #
  # First year represented by each publication
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(
      startperiod_midyear,
      na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Arrange publications and observations
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_midyear,
    endperiod_midyear
  ) %>%
  
  
  # ---------------------------------------------- #
  # Publication factor follows arranged order
  
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  ) %>%
  
  
  # ============================================== #
  # Give each reference an equal-height band
  # within each realm
  
  group_by(realm) %>%
  mutate(
    
    # reference order within realm
    reference_order = match(
      publication_id,
      unique(publication_id)
    ),
    
    # number of references in realm
    n_ref_realm = n_distinct(publication_id)
    
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Give each observation a position within
  # its reference band
  
  group_by(realm, publication_id) %>%
  mutate(
    
    n_entry = n(),
    obs_order = row_number(),
    
    # center of this reference's band
    reference_center = (
      reference_order - 0.5
    ) / n_ref_realm,
    
    # total vertical height available
    # to this reference
    reference_height = 1 / n_ref_realm,
    
    # observation positions within reference band
    #
    # observations occupy 70% of the available
    # reference band (±35% around the center)
    obs_offset = if (n() == 1) {
      
      0
      
    } else {
      
      (
        (obs_order - 1) / (n_entry - 1) - 0.5
      ) * 0.7 * first(reference_height)
      
    },
    
    # final y position
    obs_y = reference_center + obs_offset
    
  ) %>%
  
  ungroup() %>%
  
  mutate(
    realm = factor(
      realm,
      levels = c("terrestrial", "marine"),
      labels = c("Terrestrial", "Marine")
    ),
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


# ============================================== #
# Reference labels
#
# Because each reference now has its own
# equal-height band, simply use reference_center
# as the label position.


publication_labels <- rt_temp %>%
  
  group_by(
    realm,
    publication_id
  ) %>%
  
  summarise(
    label_y = first(reference_center),
    n_entry = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    publication_label = as.character(publication_id),
    
    publication_label = case_when(
      
      # References containing "et al."
      stringr::str_detect(publication_label, " et al\\. ") ~
        stringr::str_replace(
          publication_label,
          "^(.*?) et al\\. (.*)$",
          "'\\1 '*italic('et al.')*' \\2'"
        ),
      
      # References without "et al."
      TRUE ~ paste0(
        "'",
        publication_label,
        "'"
      )
    ),
    
    # Add italic (n = x)
    publication_label = paste0(
      publication_label,
      "*' ('*italic(n == ",
      n_entry,
      ")*')'"
    )
  )


# ============================================== #
# Publication colors


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


# ============================================== #
# Plot


ggplot(
  rt_temp,
  aes(
    y = obs_y,
    color = publication_id
  )
) +
  
  # ---------------------------------------------- #
  # Study period
  
  geom_segment(
    aes(
      x = startperiod_midyear,
      xend = endperiod_midyear,
      yend = obs_y,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at first survey year
  
  geom_text(
    aes(
      x = startperiod_midyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at final survey year
  
  geom_text(
    aes(
      x = endperiod_midyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Reference labels
  
  geom_text(
    data = publication_labels,
    aes(
      x = -Inf,
      y = label_y,
      label = publication_label
    ),
    inherit.aes = FALSE,
    parse = TRUE,
    hjust = 1.05,
    size = 3.5
  ) +
  
  
  # ============================================== #
  # Scales
  
  scale_linetype_manual(
    values = c(
      "2 censuses" = "longdash",
      ">2 censuses" = "solid"
    )
  ) +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  scale_x_continuous(
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  
  scale_y_reverse(
    limits = c(1.02, -0.02),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # Equal-height realm facets
  
  facet_grid2(
    rows = vars(realm),
    scales = "fixed",
    space = "fixed"
  ) +
  
  
  coord_cartesian(
    clip = "off"
  ) +
  
  
  labs(
    x = "Year of record",
    y = NULL
  ) +
  
  
  guides(
    color = "none",
    linetype = "none"
  ) +
  
  
  theme_bw(base_size = 16) +
  
  theme(
    legend.position = "none",
    
    strip.placement = "outside",
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    # remove y-axis grid lines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    panel.spacing.y = unit(0.15, "lines"),
    
    plot.margin = margin(
      t = 5.5,
      r = 5.5,
      b = 5.5,
      l = 150
    )
  )



```


### 3: row height is fixed regardless of sample size, start-start to end-end

```{r}
#| fig-width: 10
#| fig-height: 8

# ============================================== #
# Gantt chart-like plot of study duration per observation
#
# Equal vertical height for:
#   1. each realm
#   2. each reference within each realm


rt_temp <- rt_cleaned %>%
  
  filter(
    !is.na(startperiod_startyear),
    !is.na(endperiod_endyear)
  ) %>%
  
  
  # ---------------------------------------------- #
  # First year represented by each publication
  
  group_by(realm, publication_id) %>%
  mutate(
    publication_start = min(
      startperiod_startyear,
      na.rm = TRUE
    )
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Arrange publications and observations
  
  arrange(
    realm,
    publication_start,
    publication_id,
    startperiod_startyear,
    endperiod_endyear
  ) %>%
  
  
  # ---------------------------------------------- #
  # Publication factor follows arranged order
  
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  ) %>%
  
  
  # ============================================== #
  # Give each reference an equal-height band
  # within each realm
  
  group_by(realm) %>%
  mutate(
    
    # reference order within realm
    reference_order = match(
      publication_id,
      unique(publication_id)
    ),
    
    # number of references in realm
    n_ref_realm = n_distinct(publication_id)
    
  ) %>%
  ungroup() %>%
  
  
  # ---------------------------------------------- #
  # Give each observation a position within
  # its reference band
  
  group_by(realm, publication_id) %>%
  mutate(
    
    n_entry = n(),
    obs_order = row_number(),
    
    # center of this reference's band
    reference_center = (
      reference_order - 0.5
    ) / n_ref_realm,
    
    # total vertical height available
    # to this reference
    reference_height = 1 / n_ref_realm,
    
    # observation positions within reference band
    #
    # observations occupy 70% of the available
    # reference band (±35% around the center)
    obs_offset = if (n() == 1) {
      
      0
      
    } else {
      
      (
        (obs_order - 1) / (n_entry - 1) - 0.5
      ) * 0.7 * first(reference_height)
      
    },
    
    # final y position
    obs_y = reference_center + obs_offset
    
  ) %>%
  
  ungroup() %>%
  
  mutate(
    realm = factor(
      realm,
      levels = c("terrestrial", "marine"),
      labels = c("Terrestrial", "Marine")
    ),
    census_linetype = case_when(
      number_of_censuses == 2 ~ "2 censuses",
      TRUE ~ ">2 censuses"
    )
  )


# ============================================== #
# Reference labels
#
# Because each reference now has its own
# equal-height band, simply use reference_center
# as the label position.


publication_labels <- rt_temp %>%
  
  group_by(
    realm,
    publication_id
  ) %>%
  
  summarise(
    label_y = first(reference_center),
    n_entry = n(),
    .groups = "drop"
  ) %>%
  
  mutate(
    publication_label = as.character(publication_id),
    
    publication_label = case_when(
      
      # References containing "et al."
      stringr::str_detect(publication_label, " et al\\. ") ~
        stringr::str_replace(
          publication_label,
          "^(.*?) et al\\. (.*)$",
          "'\\1 '*italic('et al.')*' \\2'"
        ),
      
      # References without "et al."
      TRUE ~ paste0(
        "'",
        publication_label,
        "'"
      )
    ),
    
    # Add italic (n = x)
    publication_label = paste0(
      publication_label,
      "*' ('*italic(n == ",
      n_entry,
      ")*')'"
    )
  )


# ============================================== #
# Publication colors


n_pub <- nlevels(rt_temp$publication_id)

publication_colors <- colorspace::qualitative_hcl(
  n_pub,
  palette = "Dark 3"
)

publication_colors <- publication_colors[
  c(
    seq(1, n_pub, by = 2),
    seq(2, n_pub, by = 2)
  )
]


# ============================================== #
# Plot


ggplot(
  rt_temp,
  aes(
    y = obs_y,
    color = publication_id
  )
) +
  
  # ---------------------------------------------- #
  # Study period
  
  geom_segment(
    aes(
      x = startperiod_startyear,
      xend = endperiod_endyear,
      yend = obs_y,
      linetype = census_linetype
    ),
    linewidth = 1,
    alpha = 1
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at first survey year
  
  geom_text(
    aes(
      x = startperiod_startyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Plus sign at final survey year
  
  geom_text(
    aes(
      x = endperiod_endyear,
      y = obs_y
    ),
    label = "+",
    color = "black",
    size = 4,
    vjust = 0.35,
    show.legend = FALSE
  ) +
  
  
  # ---------------------------------------------- #
  # Reference labels
  
  geom_text(
    data = publication_labels,
    aes(
      x = -Inf,
      y = label_y,
      label = publication_label
    ),
    inherit.aes = FALSE,
    parse = TRUE,
    hjust = 1.05,
    size = 3.5
  ) +
  
  
  # ============================================== #
  # Scales
  
  scale_linetype_manual(
    values = c(
      "2 censuses" = "longdash",
      ">2 censuses" = "solid"
    )
  ) +
  
  scale_color_manual(
    values = publication_colors
  ) +
  
  scale_x_continuous(
    expand = expansion(
      mult = c(0.02, 0.02)
    )
  ) +
  
  scale_y_reverse(
    limits = c(1.02, -0.02),
    expand = c(0, 0)
  ) +
  
  
  # ============================================== #
  # Equal-height realm facets
  
  facet_grid2(
    rows = vars(realm),
    scales = "fixed",
    space = "fixed"
  ) +
  
  
  coord_cartesian(
    clip = "off"
  ) +
  
  
  labs(
    x = "Year of record",
    y = NULL
  ) +
  
  
  guides(
    color = "none",
    linetype = "none"
  ) +
  
  
  theme_bw(base_size = 16) +
  
  theme(
    legend.position = "none",
    
    strip.placement = "outside",
    strip.text.y.right = element_text(angle = 0),
    strip.background = element_rect(fill = "grey90"),
    
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    # remove y-axis grid lines
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    panel.spacing.y = unit(0.15, "lines"),
    
    plot.margin = margin(
      t = 5.5,
      r = 5.5,
      b = 5.5,
      l = 150
    )
  )



```



```{r}
#| fig-width: 12
#| fig-height: 12


# ============================================== #
# bar plot: count rows per ecosystem x gradient x position

rt_temp <- rt_cleaned %>%
  count(
    gradient,
    ecosystem,
    position,
    realm
  )


p <- ggplot(
  rt_temp,
  aes(
    x = ecosystem,
    y = n,
    fill = realm
  )
) +
  
  geom_col(position = "stack") +
  
  facet_grid2(
    position ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  
  labs(
    x = "Ecosystem",
    y = "Number of observations",
    fill = "Realm",
    title = "Number of observations per gradient per position"
  ) +
  
  theme_bw(base_size = 16) +
  
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )

suppressWarnings(print(p))
```



```{r}
#| fig-width: 10
#| fig-height: 5


# ============================================== #
# boxplot per ecosystem x magnitude x gradient

# convert to numeric (safe)
rt_temp <- rt_cleaned %>%
  filter(
    gradient != "na",
    !is.na(magnitude_of_shift)
  ) %>%
  arrange(publication_year, publication_id) %>%
  mutate(
    publication_id = factor(
      publication_id,
      levels = unique(publication_id)
    )
  )


ggplot(rt_temp, aes(x = realm, y = magnitude_of_shift)) +
  
  # horizontal reference line
  geom_hline(
    yintercept = 0,
    color = "black",
    linewidth = 0.6
  ) +
  # geom_boxplot(
  #   fill = "grey70",
  #   color = "black",
  #   outlier.shape = NA
  # ) +
  geom_jitter(
    aes(color = publication_id),
    width = 0.2,
    alpha = 0.8
  ) +
  facet_grid2(
    . ~ gradient,
    scales = "free",
    independent = "y"
  ) +
  labs(
    x = "Realm",
    y = "Magnitude of shift",
    color = "Reference no",
    title = "Shift magnitude across realms and gradients"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


```{r}
#| fig-width: 10
#| fig-height: 5

# ============================================== #
# bar plot: count rows per ecosystem x shift_tested x gradient

rt_temp <- rt_cleaned %>%
  filter(gradient != "na") %>%
  count(
    gradient,
    realm,
    shift_tested,
    realm
  )


ggplot(
  rt_temp,
  aes(
    x = realm,
    y = n,
    fill = shift_tested
  )
) +
  geom_col(position = "stack") +
  facet_grid2(
    . ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  labs(
    x = "Realm",
    y = "Number of observations",
    fill = "Shift tested"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )
```


```{r}
#| fig-width: 10
#| fig-height: 5


# ============================================== #
# bar plot: count rows per ecosystem x shift_significant x gradient

rt_temp <- rt_cleaned %>%
  filter(
    gradient != "na",
    shift_tested == "yes"
  ) %>%
  count(
    gradient,
    realm,
    shift_significant,
    realm
  )


ggplot(
  rt_temp,
  aes(
    x = realm,
    y = n,
    fill = shift_significant
  )
) +
  geom_col(position = "stack") +
  facet_grid2(
    . ~ gradient,
    scales = "free_x",
    independent = "x"
  ) +
  labs(
    x = "Realm",
    y = "Number of observations",
    fill = "Shift significant"
  ) +
  theme_bw(base_size = 16) +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    )
  )






```





# Session info

```{r}
#| label: session-info

sessioninfo::session_info()
```

```{r}
#| label: run-info
#| echo: false

time_end <- Sys.time()
elapsed_sec <- as.numeric(difftime(time_end, time_start, units = "secs"))
elapsed_min <- as.numeric(difftime(time_end, time_start, units = "mins"))

readr::write_lines(
  c(
    paste("rendered_at:", Sys.time()),
    paste("input_file:", in1),
    paste("input_file:", in2),
    paste("output_dir:", dir_out),
    paste("run_time_seconds:", round(elapsed_sec, 1)),
    paste("run_time_minutes:", round(elapsed_min, 2))),
  file.path(dir_out, "run_info.txt")
)
```
